---
title: "Hidden Markov model fitting"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(zoo)
library(here)
library(purrr)
library(scales)
library(ggforce)
library(ggspatial)
library(patchwork)
library(cowplot)
library(knitr)
library(kableExtra)
library(sf)
library(sp)
library(geosphere)
library(rnaturalearth)
library(rnaturalearthdata)
library(rosm)
library(raster)
library(ggeffects)
library(splines)
library(momentuHMM)
library(glmmTMB)

source(here::here("code", "_shared.R"))
```

```{r directories, include=FALSE}
ensure_project_dirs()
```

# 03: fit and save heavy HMM objects

Run crawlWrap and HMM fitting, then save objects needed by the downstream script as RDS files.

# Data input

Analysis settings

```{r}
HMM_SETTINGS <- list(
  section_gap_min = 30,
  section_min_points = 13, # 13 five-minute points: more than 60 min
  min_nest_distance_km = 1,
  sea_block_min_points = 6,
  random_seed = 123,
  random_starts = 20,
  individual_random_starts = 20
)

HMM_PATHS <- list(
  breakwater = here::here("rawdata", "Kab_4.gpkg"),
  coastline = here::here("rawdata", "gadm41_JPN_shp", "gadm41_JPN_0.shp"),
  gps = here::here("outputs", "resampling.csv"),
  fitted_objects = here::here("outputs", "03_HMM", "03_fit_objects.rds")
)

assert_input_files(unlist(HMM_PATHS[c("breakwater", "coastline", "gps")]))

# Keep the original variable names below so the statistical pipeline is unchanged.
cut_time <- HMM_SETTINGS$section_gap_min
min_sec_time <- HMM_SETTINGS$section_min_points
min_nest_dist <- HMM_SETTINGS$min_nest_distance_km
nest <- unname(NEST_COORDS)
```

maps

```{r}
# Breakwater polygon
kab <- st_read(HMM_PATHS$breakwater, layer = "kab4__kab")
kab$category <- ifelse(!is.na(kab$breakwater), "breakwater", NA)
b_sf <- kab[kab$category == "breakwater", ]
b_sf <- st_make_valid(b_sf) # Fix invalid geometries

# GADM
coast <- st_read(HMM_PATHS$coastline)
coast <- st_make_valid(coast)
```

GPS data loading

```{r}
# GPS data
data0 <- read_csv(HMM_PATHS$gps)

# convert to sf object
data0 <- data0 %>% st_as_sf(coords = c("lon", "lat"), crs = 4326)

# extract lon/lat and drop geometry
coords <- sf::st_coordinates(data0)
data0 <- data0 %>%
  dplyr::mutate(lon = coords[, "X"], lat = coords[, "Y"]) %>%
  sf::st_drop_geometry()
```

task score

```{r}
# success rate
score <- tribble(
  ~id, ~success_rate,
  "62206", 0.40,
  "62207", 0.00,
  "671",   0.60,
  "675",   0.50,
  "681",   0.40,
  "683",   0.00,
  "685",   0.38,
  "ATK_29B_S1", 0.00,
  "ATK_2B_S1",  0.40,
  "ATK_6E_S1",  0.60,
  "BTG_KAB_2025_9B01831_S1", 0.11,
  "BTG_KAB_2025_9B03427_S1", 0.67,
  "BTG_KAB_2025_9B40989_S2", 0.25
)


# z-standardize (mean = 0, SD = 1)
score <-score %>% mutate(score_sc = as.numeric(scale(success_rate)))

success_rate_by_id <- score %>%
  dplyr::transmute(
    id = as.character(id),
    success_rate
  )

data0 <- data0 %>% filter(id %in% score$id)

cat(length(unique(data0$id)), "individuals are included in the data.\n")

data0$id %>% unique()
```

# Data processing

exclude points within min_nest_dist of the nest

```{r}
data0 <- data0 %>%
  mutate(nest_dist  = distGeo(nest, dplyr::select(., lon, lat)) / 1000) %>% # distance from nest (km)
  filter(nest_dist >= min_nest_dist)
```

Create continuous sections and unique section identifiers

```{r}
data <- data0 %>%
  group_by(id) %>%
  arrange(time) %>% 
  mutate(
    time_diff = c(0, as.numeric(diff(time), units = "secs") / 60),  # time gap between points (min)
    section_id = cumsum(time_diff > cut_time) + 1                   # start new section if gap > cut_time
  ) %>%
  group_by(id, section_id) %>%
  filter(n() >= min_sec_time) %>%                                   
  ungroup() %>%
  mutate(ID = paste(id, section_id, sep = "_")) %>%                 # unique section ID
  dplyr::select(-time_diff)
```

estuary classification

Identify whether each GPS location was recorded within the estuary area, defined as being within 800 m of estuary_1 or 500 m of estuary_2

```{r}
#* Define the coordinates of the two estuary reference points
estuary_1 <- c(141.508143, 40.543210)
estuary_2 <- c(141.462772, 40.601823)

#* Define the radius around each estuary point (in meters)
radius1_m <- 0.8 * 1000  # km -> m
radius2_m <- 0.5 * 1000

#* Classify each location as estuary-associated if it falls within
#* the specified radius of either estuary reference point
data <- data %>%rowwise() %>% 
  mutate(estuary = (distHaversine(c(lon, lat), estuary_1) <= radius1_m) | (distHaversine(c(lon, lat), estuary_2) <= radius2_m)
  ) %>%
  ungroup()
```

specify time zone with NOAA NCAT

```{r}
rawData <- dplyr::select(data, time, lon, lat, ID, section_id, estuary) 

llcoord  <- SpatialPoints(rawData[,2:3], proj4string=CRS("+proj=longlat +datum=WGS84"))
utmcoord <- spTransform(llcoord, CRS("+proj=utm +zone=54 ellps=WGS84"))

rawData$x <- attr(utmcoord, "coords")[,1]
rawData$y <- attr(utmcoord, "coords")[,2]
```

estuary classification 2

```{r}
rawData <- rawData %>%
  mutate(id = sub("_([^_]+)$", "", ID)) %>%
  arrange(id, time) %>%
  group_by(id) %>%
  mutate(section_change = (section_id != lag(section_id, default = first(section_id))),   
    after_estuary = (estuary == FALSE & lag(estuary, default = FALSE) == TRUE),
    new_section_flag = section_change | after_estuary,
    section_id_est = cumsum(new_section_flag) + 1,
    ID_est = paste0(id, "_", section_id_est)
  ) %>%
  ungroup() %>% dplyr::select(-section_change, -after_estuary, -new_section_flag)

rawData <- rawData %>% arrange(ID, time)
```

# crawlWrap → prepData

```{r}
set.seed(HMM_SETTINGS$random_seed)

# crawlWrap one ID at a time, retrying failed IDs automatically
# Final output: df_crw


# Convert crwPredict object to a plain data.frame
as_plain_crw_df <- function(x) {
  x <- as.data.frame(x)
  class(x) <- setdiff(class(x), "crwPredict")
  x
}

# Run crawlWrap once for a single ID with a given setting
run_crawlwrap_once <- function(one, setting, time_step = "5 min") {
  
  args <- list(
    obsData   = one,
    timeStep  = time_step,
    fixPar    = c(NA, NA),
    attempts  = setting$attempts,
    retryFits = setting$retryFits,
    retrySD   = setting$retrySD
  )
  
  # Important:
  # If setting$theta is NULL, do not pass theta at all.
  # Passing theta = NULL explicitly can cause an error in crawlWrap.
  if (!is.null(setting$theta)) {
    args$theta <- setting$theta
  }
  
  do.call(momentuHMM::crawlWrap, args)
}

# Settings to try for each ID.
# The first setting is your original setting.
# Later settings are only used if the ID fails.
crawl_retry_plan <- list(
  list(
    label     = "original_theta_10_-1.5_sd1",
    theta     = c(10, -1.5),
    attempts  = 100,
    retryFits = 10,
    retrySD   = 1
  ),
  list(
    label     = "theta_10_-1.5_more_attempts_sd1",
    theta     = c(10, -1.5),
    attempts  = 300,
    retryFits = 20,
    retrySD   = 1
  ),
  list(
    label     = "theta_10_-1.5_sd0.5",
    theta     = c(10, -1.5),
    attempts  = 300,
    retryFits = 20,
    retrySD   = 0.5
  ),
  list(
    label     = "theta_10_-1.5_sd2",
    theta     = c(10, -1.5),
    attempts  = 300,
    retryFits = 20,
    retrySD   = 2
  ),
  list(
    label     = "theta_6.855_-0.007_sd1",
    theta     = c(6.855, -0.007),
    attempts  = 300,
    retryFits = 20,
    retrySD   = 1
  ),
  list(
    label     = "theta_0_0_sd1",
    theta     = c(0, 0),
    attempts  = 300,
    retryFits = 20,
    retrySD   = 1
  ),
  list(
    label     = "theta_missing_default_sd1",
    theta     = NULL,   # this means theta is omitted from the crawlWrap call
    attempts  = 300,
    retryFits = 20,
    retrySD   = 1
  )
)

# Fit one ID until success, or until all settings/cycles are exhausted
fit_one_id_until_success <- function(
  target_id,
  dat,
  retry_plan = crawl_retry_plan,
  max_cycles = 5,
  time_step = "5 min"
) {
  
  one <- dat %>%
    dplyr::filter(ID == target_id) %>%
    dplyr::arrange(time) %>%
    as.data.frame()
  
  if (nrow(one) < 3) {
    return(list(
      success = FALSE,
      ID = target_id,
      pred = NULL,
      log = tibble::tibble(
        ID = target_id,
        cycle = NA_integer_,
        setting = "too_few_rows",
        success = FALSE,
        n_pred = 0L,
        error = "Fewer than 3 rows",
        n_warnings = 0L
      )
    ))
  }
  
  logs <- list()
  log_i <- 1
  
  for (cycle in seq_len(max_cycles)) {
    
    for (k in seq_along(retry_plan)) {
      
      setting <- retry_plan[[k]]
      
      message(
        "Fitting ID: ", target_id,
        " | cycle: ", cycle,
        " | setting: ", setting$label
      )
      
      warnings_here <- character()
      
      fit_try <- tryCatch(
        withCallingHandlers(
          run_crawlwrap_once(
            one = one,
            setting = setting,
            time_step = time_step
          ),
          warning = function(w) {
            warnings_here <<- c(warnings_here, conditionMessage(w))
            invokeRestart("muffleWarning")
          }
        ),
        error = function(e) e
      )
      
      # If crawlWrap itself failed
      if (inherits(fit_try, "error")) {
        logs[[log_i]] <- tibble::tibble(
          ID = target_id,
          cycle = cycle,
          setting = setting$label,
          success = FALSE,
          n_pred = 0L,
          error = conditionMessage(fit_try),
          n_warnings = length(warnings_here)
        )
        log_i <- log_i + 1
        next
      }
      
      # If crawlWrap returned no crwPredict
      if (!"crwPredict" %in% names(fit_try) || is.null(fit_try$crwPredict)) {
        logs[[log_i]] <- tibble::tibble(
          ID = target_id,
          cycle = cycle,
          setting = setting$label,
          success = FALSE,
          n_pred = 0L,
          error = "No crwPredict returned",
          n_warnings = length(warnings_here)
        )
        log_i <- log_i + 1
        next
      }
      
      pred <- as_plain_crw_df(fit_try$crwPredict)
      
      # Force ID to remain the original section ID.
      # This is critical for keeping the downstream ID/ID_est logic unchanged.
      pred$ID <- as.character(target_id)
      
      # Basic validity check
      required_cols <- c("ID", "time", "mu.x", "mu.y", "estuary", "ID_est")
      pred_ok <- nrow(pred) > 0 &&
        all(required_cols %in% names(pred)) &&
        all(!is.na(pred$ID)) &&
        any(!is.na(pred$mu.x)) &&
        any(!is.na(pred$mu.y))
      
      if (!pred_ok) {
        logs[[log_i]] <- tibble::tibble(
          ID = target_id,
          cycle = cycle,
          setting = setting$label,
          success = FALSE,
          n_pred = nrow(pred),
          error = "crwPredict returned, but required columns or mu.x/mu.y are missing",
          n_warnings = length(warnings_here)
        )
        log_i <- log_i + 1
        next
      }
      
      logs[[log_i]] <- tibble::tibble(
        ID = target_id,
        cycle = cycle,
        setting = setting$label,
        success = TRUE,
        n_pred = nrow(pred),
        error = NA_character_,
        n_warnings = length(warnings_here)
      )
      
      return(list(
        success = TRUE,
        ID = target_id,
        pred = pred,
        log = dplyr::bind_rows(logs)
      ))
    }
  }
  
  # If all attempts failed
  list(
    success = FALSE,
    ID = target_id,
    pred = NULL,
    log = dplyr::bind_rows(logs)
  )
}

# Use the same ID order as rawData
all_ids <- unique(as.character(rawData$ID))

# Run crawlWrap one ID at a time
crawl_results <- purrr::map(
  all_ids,
  ~ fit_one_id_until_success(
    target_id = .x,
    dat = rawData,
    retry_plan = crawl_retry_plan,
    max_cycles = 5,
    time_step = "5 min"
  )
)

# Attempt log
crawl_log <- dplyr::bind_rows(purrr::map(crawl_results, "log"))

# Check failed IDs
success_ids <- crawl_results %>%
  purrr::keep(~ .x$success) %>%
  purrr::map_chr("ID")

failed_ids <- setdiff(all_ids, success_ids)

cat("Original IDs:", length(all_ids), "\n")
cat("Successful IDs:", length(success_ids), "\n")
cat("Failed IDs:", length(failed_ids), "\n")

if (length(failed_ids) > 0) {
  print(failed_ids)
  print(crawl_log %>% dplyr::filter(ID %in% failed_ids))
  
  partial_rds_dir <- here::here("outputs", "HMM_YG2")
  dir.create(partial_rds_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(
    list(
      all_ids = all_ids,
      success_ids = success_ids,
      failed_ids = failed_ids,
      crawl_results = crawl_results,
      crawl_log = crawl_log,
      rawData = rawData,
      score = score,
      success_rate_by_id = success_rate_by_id
    ),
    file.path(partial_rds_dir, "HMM_YG2_crawlwrap_partial_results.rds")
  )
  cat(
    "Saved partial crawlWrap results to:",
    file.path(partial_rds_dir, "HMM_YG2_crawlwrap_partial_results.rds"),
    "\n"
  )
  
  stop("Some IDs still failed in crawlWrap. Check crawl_log.")
}

# Combine all successful predictions.
# This becomes the same downstream object as before: df_crw
df_crw <- crawl_results %>%
  purrr::map("pred") %>%
  dplyr::bind_rows() %>%
  dplyr::mutate(
    ID = as.character(ID),
    time = as.POSIXct(time, tz = "UTC")
  ) %>%
  dplyr::arrange(ID, time) %>%
  as.data.frame()

# Same downstream processing as your original code
df_crw <- df_crw %>%
  dplyr::mutate(id = sub("_[^_]+$", "", ID))

df_crw <- df_crw %>%
  dplyr::left_join(score %>% dplyr::distinct(id, score_sc), by = "id")

# Final checks
df_crw_check <- df_crw %>%
  dplyr::summarise(
    n = dplyr::n(),
    n_ID = dplyr::n_distinct(ID),
    NA_ID = sum(is.na(ID)),
    NA_time = sum(is.na(time)),
    NA_mu.x = sum(is.na(mu.x)),
    NA_mu.y = sum(is.na(mu.y)),
    NA_estuary = sum(is.na(estuary)),
    NA_ID_est = sum(is.na(ID_est)),
    NA_score_sc = sum(is.na(score_sc))
  )

print(df_crw_check)


critical_na_cols <- c("NA_ID", "NA_time", "NA_mu.x", "NA_mu.y", "NA_score_sc")
critical_na <- unlist(df_crw_check[critical_na_cols])
if (any(critical_na > 0)) {
  stop(
    "Critical NA values remain after crawlWrap: ",
    paste(names(critical_na)[critical_na > 0], critical_na[critical_na > 0], sep = "=", collapse = ", ")
  )
}

if (df_crw_check$NA_estuary > 0) {
  warning("NA values remain in estuary before locf: ", df_crw_check$NA_estuary)
}

if (df_crw_check$NA_ID_est > 0) {
  warning("NA values remain in ID_est before locf: ", df_crw_check$NA_ID_est)
}

# Optional: inspect which settings succeeded
crawl_log %>%
  dplyr::filter(success) %>%
  dplyr::count(setting)
```

Prepare sea-track data for HMM

```{r}
# Fill estuary labels within each individual
df_crw2 <- df_crw %>%
  dplyr::arrange(ID, time) %>%
  dplyr::group_by(ID) %>%
  dplyr::mutate(
    estuary = zoo::na.locf(zoo::na.locf(estuary, na.rm = FALSE), fromLast = TRUE, na.rm = FALSE),
    ID_est = zoo::na.locf(zoo::na.locf(ID_est, na.rm = FALSE), fromLast = TRUE, na.rm = FALSE)
  ) %>%
  dplyr::ungroup() %>%
  as.data.frame()

df_crw2_check <- df_crw2 %>%
  dplyr::summarise(
    NA_estuary = sum(is.na(estuary)),
    NA_ID_est = sum(is.na(ID_est))
  )

print(df_crw2_check)

if (df_crw2_check$NA_estuary > 0 || df_crw2_check$NA_ID_est > 0) {
  stop(
    "NA values remain after filling estuary/ID_est labels: ",
    paste(names(df_crw2_check), unlist(df_crw2_check), sep = "=", collapse = ", ")
  )
}


# Remove estuary points and short ID_est segments
df_keep <- df_crw2 %>%
  dplyr::arrange(ID, time) %>%
  dplyr::filter(estuary == FALSE) %>%
  dplyr::group_by(ID_est) %>%
  dplyr::filter(dplyr::n() >= 6) %>%
  dplyr::ungroup() %>%
  as.data.frame()


# Classify locations as land, sea, or breakwater
data_xy <- sf::st_as_sf(
  df_keep,
  coords = c("mu.x", "mu.y"),
  crs = 32654
)

data_xy_ll <- sf::st_transform(data_xy, 4326)
coast_wgs <- sf::st_transform(coast, 4326)

sf::sf_use_s2(TRUE)

data_located <- sf::st_join(data_xy_ll, coast_wgs, left = TRUE) %>%
  dplyr::mutate(
    coast = ifelse(is.na(COUNTRY), "sea", "land")
  )

sf::sf_use_s2(FALSE)

breakwater_hit <- lengths(sf::st_intersects(data_located, b_sf)) > 0

data_located <- data_located %>%
  dplyr::mutate(
    on_breakwater = breakwater_hit,
    location = ifelse(on_breakwater, "land", coast)
  ) %>%
  dplyr::select(-dplyr::any_of(c("GID_0", "COUNTRY"))) %>%
  sf::st_drop_geometry() %>%
  as.data.frame()

df_keep <- df_keep %>%
  dplyr::left_join(
    data_located %>%
      dplyr::select(ID_est, time, location, on_breakwater),
    by = c("ID_est", "time")
  )


# Rename ID columns for HMM analysis
df_keep <- df_keep %>%
  dplyr::select(-dplyr::any_of(c("x", "y"))) %>%
  dplyr::rename(
    ID_real = ID,
    ID = ID_est
  ) %>%
  dplyr::mutate(
    ID = as.character(ID),
    ID_real = as.character(ID_real),
    time = as.POSIXct(time, tz = "UTC"),
    mu.x = as.numeric(mu.x),
    mu.y = as.numeric(mu.y)
  ) %>%
  as.data.frame()


# Identify continuous sea/land blocks
df_keep <- df_keep %>%
  dplyr::arrange(ID, time) %>%
  dplyr::group_by(ID) %>%
  dplyr::mutate(
    loc_bin = ifelse(location == "land", "land", "sea"),
    block_change = loc_bin != dplyr::lag(loc_bin, default = dplyr::first(loc_bin)),
    block_num = cumsum(block_change),
    ID_block = paste0(ID, "_", loc_bin, "_", block_num)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-block_change, -block_num)


# Keep only sea blocks that meet the pre-specified minimum length.
df_sea <- df_keep %>%
  dplyr::filter(loc_bin == "sea") %>%
  dplyr::group_by(ID_block) %>%
  dplyr::filter(dplyr::n() >= HMM_SETTINGS$sea_block_min_points) %>%
  dplyr::ungroup()

cat("Number of sea blocks:", dplyr::n_distinct(df_sea$ID_block), "\n")


# Prepare data for momentuHMM
cols_needed <- c(
  "ID_block",
  "time",
  "mu.x",
  "mu.y",
  "location",
  "on_breakwater",
  "id",
  "ID_real",
  "loc_bin"
)

Data_pre <- momentuHMM::prepData(
  data = df_sea %>%
    dplyr::select(dplyr::all_of(cols_needed)) %>%
    dplyr::rename(ID = ID_block) %>%
    as.data.frame(),
  coordNames = c("mu.x", "mu.y")
)


# Add cognitive score
Data_HMM <- Data_pre

Data_HMM$id <- sub("_[^_]+$", "", Data_HMM$ID_real)
Data_HMM$score_sc <- score$score_sc[match(Data_HMM$id, score$id)]
Data_HMM$id_factor <- factor(Data_HMM$id)

# Replace zero step lengths with a small positive value
Data_HMM$step[
  !is.na(Data_HMM$step) & Data_HMM$step == 0
] <- 1e-6
```

# HMM with covariates

HMM parameters

```{r}
dist <- list(step = "gamma", angle = "vm")

jitter_Par0_state <- function(Par0, prop = 0.2){
  lapply(Par0, function(x){
    is_coef0 <- x == 0
    out <- x
    out[!is_coef0] <- x[!is_coef0] + log(runif(sum(!is_coef0), 1 - prop, 1 + prop))
    out
  })
}

stateNames <- HMM_STATE_NAMES

DM <- list(
  step = list(
    mean = ~ state1(score_sc) + state2(score_sc),
    sd   = ~ 1
  ),
  angle = list(
    concentration = ~ state1(score_sc) + state2(score_sc)
  )
)

step_mean   <- c(2600, 900, 50, 100)
step_sd     <- c(800, 500, 50, 50)
angle_conc  <- c(10, 1, 0.1, 8)

Par0 <- list(
  step = c(
    log(step_mean[1]), 0,
    log(step_mean[2]), 0,
    log(step_mean[3]),
    log(step_mean[4]),
    log(step_sd[1]),
    log(step_sd[2]),
    log(step_sd[3]),
    log(step_sd[4])
  ),
  angle = c(
    log(angle_conc[1]), 0,
    log(angle_conc[2]), 0,
    log(angle_conc[3]), 
    log(angle_conc[4])
  )
)
```

fitHMM

```{r}
set.seed(HMM_SETTINGS$random_seed)

n_try <- HMM_SETTINGS$random_starts
model_list <- list()
aic_values <- rep(NA_real_, n_try)
conv <- rep(FALSE, n_try)
init_list <- vector("list", n_try)

for(i in 1:n_try){
  if(i == 1){
    Par0_try <- Par0
  } else {
    Par0_try <- jitter_Par0_state(Par0, prop = 0.2)
  }
  
  init_list[[i]] <- Par0_try
  
  tryCatch({
    m <- fitHMM(
      data = Data_HMM,
      nbStates = 4,
      dist = dist,
      Par0 = Par0_try,
      DM = DM,
      formula = ~ score_sc,
      estAngleMean = list(angle = FALSE),
      stateNames = stateNames
    )
    
    model_list[[i]] <- m
    aic_values[i] <- AIC(m)
    conv[i] <- TRUE
    
    cat("model", i, ": AIC =", aic_values[i], "\n")
  }, error = function(e){
    cat("model", i, "failed:", e$message, "\n")
    conv[i] <<- FALSE
    aic_values[i] <<- NA_real_
  })
}

init_summary <- do.call(rbind, lapply(seq_along(init_list), function(i){
  data.frame(
    run = i,
    step_1 = init_list[[i]]$step[1],
    step_2 = init_list[[i]]$step[2],
    step_3 = init_list[[i]]$step[3],
    step_4 = init_list[[i]]$step[4],
    step_5 = init_list[[i]]$step[5],
    step_6 = init_list[[i]]$step[6],
    step_7 = init_list[[i]]$step[7],
    step_8 = init_list[[i]]$step[8],
    step_9 = init_list[[i]]$step[9],
    step_10 = init_list[[i]]$step[10],
    angle_1 = init_list[[i]]$angle[1],
    angle_2 = init_list[[i]]$angle[2],
    angle_3 = init_list[[i]]$angle[3],
    angle_4 = init_list[[i]]$angle[4],
    angle_5 = init_list[[i]]$angle[5],
    angle_6 = init_list[[i]]$angle[6]
  )
}))

summary <- data.frame(
  run = 1:n_try,
  AIC = aic_values,
  converged = conv
) %>%
  dplyr::left_join(init_summary, by = "run")

print(summary)

summary %>%
  dplyr::mutate(
    step_mean_directed_exp   = exp(step_1),
    step_mean_undirected_exp = exp(step_3),
    step_mean_ARS_exp        = exp(step_5),
    step_mean_water_exp      = exp(step_6)
  ) %>%
  dplyr::select(
    run,
    AIC,
    converged,
    step_mean_directed_exp,
    step_mean_undirected_exp,
    step_mean_ARS_exp,
    step_mean_water_exp
  )
```

Summary of the best HMM

```{r}
results_summary <- data.frame(
  run       = 1:n_try,
  AIC       = aic_values,
  converged = conv
)
print(results_summary)

valid_runs <- which(conv & !is.na(aic_values))
if (length(valid_runs) == 0) {
  stop("No sea models converged successfully.")
}

best_run <- valid_runs[which.min(aic_values[valid_runs])]
cat("Best run:", best_run, "\n")
cat("Minimum AIC:", aic_values[best_run], "\n")

best <- model_list[[best_run]]
CIbeta(best)
best
```

 State-dependent HMM parameter estimates at score_sc = 0

```{r}
score0 <- 0

step_beta<- unlist(best$mle$step)
angle_beta <- unlist(best$mle$angle)

param_table <- data.frame(
  state = stateNames,
  step_mean = c(
    exp(step_beta[1] + step_beta[2] * score0),
    exp(step_beta[3] + step_beta[4] * score0),
    exp(step_beta[5]),
    exp(step_beta[6])
  ),
  step_sd = exp(step_beta[7:10]),
  angle_concentration = c(
    exp(angle_beta[1] + angle_beta[2] * score0),
    exp(angle_beta[3] + angle_beta[4] * score0),
    exp(angle_beta[5]),
    exp(angle_beta[6])
  )
)

param_table <- param_table %>%
  dplyr::mutate(
    step_mean = round(step_mean, 2),
    step_sd = round(step_sd, 2),
    angle_concentration = round(angle_concentration, 3)
  )

print(param_table)
```

Extract beta, SE, and 95% CI from covariate HMM

```{r}
ci_beta <- momentuHMM::CIbeta(best)

extract_hmm_beta <- function(ci_beta, component, term, label) {
  
  tibble::tibble(
    parameter = label,
    beta = ci_beta[[component]]$est[1, term],
    SE = ci_beta[[component]]$se[1, term],
    CI_lower = ci_beta[[component]]$lower[1, term],
    CI_upper = ci_beta[[component]]$upper[1, term]
  )
}

hmm_beta_table <- dplyr::bind_rows(
  extract_hmm_beta(
    ci_beta,
    component = "step",
    term = "mean_1:score_sc",
    label = "Step length: directed flight"
  ),
  extract_hmm_beta(
    ci_beta,
    component = "step",
    term = "mean_2:score_sc",
    label = "Step length: undirected flight"
  ),
  extract_hmm_beta(
    ci_beta,
    component = "angle",
    term = "concentration_1:score_sc",
    label = "Angle concentration: directed flight"
  ),
  extract_hmm_beta(
    ci_beta,
    component = "angle",
    term = "concentration_2:score_sc",
    label = "Angle concentration: undirected flight"
  )
) %>%
  dplyr::mutate(
    dplyr::across(
      c(beta, SE, CI_lower, CI_upper),
      ~ round(.x, 3)
    )
  )

hmm_beta_table
```


Viterbi state 

```{r}
Data_HMM <- Data_HMM %>% 
  dplyr::mutate(
    ID = factor(ID),
    id = factor(id),
    state = factor(
      momentuHMM::viterbi(best),
      levels = 1:4,
      labels = stateNames
    )
  )

print(table(Data_HMM$state))
print(table(Data_HMM$state) / nrow(Data_HMM))
```


# Individual-level HMM

```{r}
stateNames_ind <- stateNames

dist_ind <- list(step = "gamma", angle = "vm")

DM_ind <- list(
  step = list(
    mean = ~ 1,
    sd   = ~ 1
  ),
  angle = list(
    concentration = ~ 1
  )
)


best_step_coef <- unlist(best$mle$step)
best_angle_coef <- unlist(best$mle$angle)

best_coef_index <- list(
  step_mean = c(directed = 1, undirected = 3, ARS = 5, water = 6),
  step_sd = c(directed = 7, undirected = 8, ARS = 9, water = 10),
  angle_conc = c(directed = 1, undirected = 3, ARS = 5, water = 6)
)

if (length(best_step_coef) < max(c(best_coef_index$step_mean, best_coef_index$step_sd))) {
  stop("best$mle$step does not contain the expected coefficients for individual-level initial values.")
}

if (length(best_angle_coef) < max(best_coef_index$angle_conc)) {
  stop("best$mle$angle does not contain the expected coefficients for individual-level initial values.")
}

Par_best_ind <- list(
  step = unname(c(
    best_step_coef[best_coef_index$step_mean],
    best_step_coef[best_coef_index$step_sd]
  )),
  angle = unname(best_angle_coef[best_coef_index$angle_conc])
)



ids <- unique(Data_HMM$id)

n_try_ind <- HMM_SETTINGS$individual_random_starts

model_by_id <- vector("list", length(ids))
aic_by_id <- rep(NA_real_, length(ids))
conv_by_id <- rep(FALSE, length(ids))
n_by_id <- rep(NA_integer_, length(ids))
best_run_by_id <- rep(NA_integer_, length(ids))
n_success_by_id <- rep(0L, length(ids))
id_model_attempts <- vector("list", length(ids))

names(model_by_id) <- as.character(ids)
names(id_model_attempts) <- as.character(ids)

for(i in seq_along(ids)){
  
  id_i <- ids[i]
  
  dat_i <- Data_HMM %>%
    dplyr::filter(id == id_i)
  
  n_by_id[i] <- nrow(dat_i)
  
  cat("\n============================\n")
  cat("id:", as.character(id_i), "\n")
  cat("n =", nrow(dat_i), "\n")
  
  model_list_i <- vector("list", n_try_ind)
  aic_i <- rep(NA_real_, n_try_ind)
  conv_i <- rep(FALSE, n_try_ind)
  error_i <- rep(NA_character_, n_try_ind)
  init_ind_list_i <- vector("list", n_try_ind)
  
  for(j in seq_len(n_try_ind)){
    
    if(j == 1){
      Par0_ind_try <- Par_best_ind
    } else {
      Par0_ind_try <- jitter_Par0_state(Par_best_ind, prop = 0.2)
    }
    
    init_ind_list_i[[j]] <- Par0_ind_try
    
    cat("individual model run", j, "/", n_try_ind, "\n")
    
    tryCatch({
      
      m_i <- momentuHMM::fitHMM(
        data = dat_i,
        nbStates = 4,
        dist = dist_ind,
        Par0 = Par0_ind_try,
        DM = DM_ind,
        formula = ~ 1,
        estAngleMean = list(angle = FALSE),
        stateNames = stateNames_ind
      )
      
      model_list_i[[j]] <- m_i
      aic_i[j] <- AIC(m_i)
      conv_i[j] <- TRUE
      
      cat("AIC =", aic_i[j], "\n")
      
    }, error = function(e){
      
      cat("failed:", e$message, "\n")
      conv_i[j] <<- FALSE
      aic_i[j] <<- NA_real_
      error_i[j] <<- e$message
      
    })
  }
  
  id_model_attempts[[as.character(id_i)]] <- tibble::tibble(
    id = as.character(id_i),
    run = seq_len(n_try_ind),
    AIC = aic_i,
    converged = conv_i,
    error = error_i
  )
  
  valid_i <- which(conv_i & !is.na(aic_i))
  n_success_by_id[i] <- length(valid_i)
  
  if(length(valid_i) > 0){
    best_j <- valid_i[which.min(aic_i[valid_i])]
    
    model_by_id[[as.character(id_i)]] <- model_list_i[[best_j]]
    aic_by_id[i] <- aic_i[best_j]
    conv_by_id[i] <- TRUE
    best_run_by_id[i] <- best_j
    
    cat("Best individual run:", best_j, "AIC =", aic_by_id[i], "\n")
  } else {
    cat("all individual HMM runs failed for id:", as.character(id_i), "\n")
    conv_by_id[i] <- FALSE
    aic_by_id[i] <- NA_real_
  }
}

id_model_attempts <- dplyr::bind_rows(id_model_attempts)

id_model_summary <- data.frame(
  id = ids,
  n = n_by_id,
  best_run = best_run_by_id,
  n_success = n_success_by_id,
  AIC = aic_by_id,
  converged = conv_by_id
)

id_model_summary
```

Viterbi state

```{r}
state_prop_by_id <- lapply(names(model_by_id), function(id_i){
  
  m_i <- model_by_id[[id_i]]
  
  if(is.null(m_i)) return(NULL)
  
  dat_i <- m_i$data %>% 
    dplyr::mutate(
      state = factor(
        momentuHMM::viterbi(m_i),
        levels = 1:4,
        labels = stateNames_ind
      )
    )
  
  dat_i %>% 
    dplyr::count(state) %>% 
    dplyr::mutate(
      id = id_i,
      prop = n / sum(n)
    ) %>% 
    dplyr::select(id, state, n, prop)
  
}) %>% 
  dplyr::bind_rows()

print(state_prop_by_id)
```

Summary of individual-level HMM parameter estimates

```{r}
extract_ind_hmm_par <- function(m_i, id_i){
  if(is.null(m_i)) return(NULL)
  nbStates_i <- m_i$conditions$nbStates
  
  # step parameters
  step_i <- m_i$mle$step
  
  if(is.matrix(step_i)){
    step_mean_i <- step_i["mean", stateNames]
    step_sd_i <- step_i["sd", stateNames]
  } else {
    step_mean_i <- step_i[1:nbStates_i]
    step_sd_i <- step_i[(nbStates_i + 1):(2 * nbStates_i)]
    names(step_mean_i) <- stateNames
    names(step_sd_i) <- stateNames
  }
  
  # angle concentration
  angle_i <- m_i$mle$angle
  
  if(is.matrix(angle_i)){
    angle_conc_i <- angle_i["concentration", stateNames]
  } else {
    if(length(angle_i) == 2 * nbStates_i){
      angle_conc_i <- angle_i[seq(2, 2 * nbStates_i, by = 2)]
    } else if(length(angle_i) == nbStates_i){
      angle_conc_i <- angle_i
    } else {
      stop(paste0("Unexpected length of angle parameters for id ", id_i, ": ", length(angle_i)))
    }
    names(angle_conc_i) <- stateNames
  }
  
  # self-transition probability
  gamma_i <- momentuHMM::getTrProbs(m_i)
  
  if(length(dim(gamma_i)) == 3){
    gamma_i <- gamma_i[, , 1]
  }
  
  rownames(gamma_i) <- stateNames
  colnames(gamma_i) <- stateNames
  
  # table
  data.frame(
    id = as.character(id_i),
    
    step_mean_directed = step_mean_i["directed"],
    step_mean_undirected = step_mean_i["undirected"],
    step_mean_ARS = step_mean_i["ARS"],
    step_mean_water = step_mean_i["water"],
    
    step_sd_directed = step_sd_i["directed"],
    step_sd_undirected = step_sd_i["undirected"],
    step_sd_ARS = step_sd_i["ARS"],
    step_sd_water = step_sd_i["water"],
    
    angle_conc_directed = angle_conc_i["directed"],
    angle_conc_undirected = angle_conc_i["undirected"],
    angle_conc_ARS = angle_conc_i["ARS"],
    angle_conc_water = angle_conc_i["water"],
    
    self_p_directed = gamma_i["directed", "directed"],
    self_p_undirected = gamma_i["undirected", "undirected"],
    self_p_ARS = gamma_i["ARS", "ARS"],
    self_p_water = gamma_i["water", "water"]
  )
}

par_by_id <- lapply(names(model_by_id), function(id_i){
  extract_ind_hmm_par(
    m_i = model_by_id[[id_i]],
    id_i = id_i
  )
}) %>%
  dplyr::bind_rows()

if (nrow(par_by_id) == 0) {
  stop("No individual-level HMM models converged successfully.")
}

id_model_summary2 <- id_model_summary %>%
  dplyr::mutate(id = as.character(id))

par_by_id <- par_by_id %>%
  dplyr::left_join(id_model_summary2, by = "id") %>%
  dplyr::select(
    id, n,
    step_mean_directed,
    step_mean_undirected,
    step_mean_ARS,
    step_mean_water,
    step_sd_directed,
    step_sd_undirected,
    step_sd_ARS,
    step_sd_water,
    angle_conc_directed,
    angle_conc_undirected,
    angle_conc_ARS,
    angle_conc_water,
    self_p_directed,
    self_p_undirected,
    self_p_ARS,
    self_p_water
  )

par_by_id
```

# Save fitted objects

```{r}
out_rds_dir <- dirname(HMM_PATHS$fitted_objects)
dir.create(out_rds_dir, recursive = TRUE, showWarnings = FALSE)

hmm_yg2_objects <- list(
  score = score,
  success_rate_by_id = success_rate_by_id,
  nest = nest,
  kab = kab,
  b_sf = b_sf,
  coast = coast,
  data0 = data0,
  data = data,
  rawData = rawData,
  crawl_results = crawl_results,
  crawl_log = crawl_log,
  df_crw_check = df_crw_check,
  df_crw2_check = df_crw2_check,
  df_crw = df_crw,
  df_crw2 = df_crw2,
  df_keep = df_keep,
  df_sea = df_sea,
  Data_pre = Data_pre,
  Data_HMM = Data_HMM,
  stateNames = stateNames,
  stateNames_ind = stateNames_ind,
  n_try = n_try,
  n_try_ind = n_try_ind,
  best = best,
  model_list = model_list,
  aic_values = aic_values,
  conv = conv,
  results_summary = results_summary,
  param_table = param_table,
  model_by_id = model_by_id,
  id_model_attempts = id_model_attempts,
  id_model_summary = id_model_summary,
  state_prop_by_id = state_prop_by_id,
  par_by_id = par_by_id
)

saveRDS(hmm_yg2_objects, HMM_PATHS$fitted_objects)
cat("Saved fitted objects to:", HMM_PATHS$fitted_objects, "\n")
```
