---
title: "HMM figures and post-hoc analyses"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(zoo)
library(here)
library(purrr)
library(scales)
library(ggforce)
library(ggspatial)
library(patchwork)
library(cowplot)
library(knitr)
library(kableExtra)
library(sf)
library(sp)
library(geosphere)
library(rnaturalearth)
library(rnaturalearthdata)
library(rosm)
library(raster)
library(ggeffects)
library(splines)
library(momentuHMM)
library(glmmTMB)
library(betareg)
library(broom)

source(here::here("code", "_shared.R"))
```

```{r directories, include=FALSE}
ensure_project_dirs()
```

# 04: figures and post-HMM analyses

Create figures and post-hoc analyses from the HMM results saved by '03_HMM_fit_and_save.Rmd'.

# Load fitted HMM objects

```{r}
fit_rds <- here::here("outputs", "03_HMM", "03_fit_objects.rds")

if (!file.exists(fit_rds)) {
  stop("03_fit_objects.rds was not found. Please run the HMM fitting and saving script first.")
}

hmm_yg2_objects <- readRDS(fit_rds)
list2env(hmm_yg2_objects, envir = .GlobalEnv)

hmm_step_minutes <- Data_HMM %>%
  dplyr::mutate(id = as.character(id)) %>%
  dplyr::arrange(id, time) %>%
  dplyr::group_by(id) %>%
  dplyr::mutate(dt_min = as.numeric(difftime(time, dplyr::lag(time), units = "mins"))) %>%
  dplyr::ungroup() %>%
  dplyr::filter(is.finite(dt_min), dt_min > 0) %>%
  dplyr::summarise(step_minutes = stats::median(dt_min, na.rm = TRUE)) %>%
  dplyr::pull(step_minutes)

calc_dwell_time_min <- function(self_transition_prob, step_minutes = hmm_step_minutes){
  ifelse(
    is.na(self_transition_prob) | self_transition_prob < 0 | self_transition_prob >= 1,
    NA_real_,
    step_minutes / (1 - self_transition_prob)
  )
}
```

### Figure S1: State-dependent HMM parameter

Distribution plot

```{r}
bg_step <- Data_HMM %>%
  dplyr::select(step) %>%
  dplyr::filter(!is.na(step))

bg_angle <- Data_HMM %>%
  dplyr::select(angle) %>%
  dplyr::filter(!is.na(angle))

get_params <- function(model){
  step_beta <- unlist(model$mle$step)
  angle_beta <- unlist(model$mle$angle)
  
  data.frame(
    state = stateNames,
    step_mean = c(
      exp(step_beta[1]),
      exp(step_beta[3]),
      exp(step_beta[5]),
      exp(step_beta[6])
    ),
    step_sd = exp(step_beta[7:10]),
    angle_conc = c(
      exp(angle_beta[1]),
      exp(angle_beta[3]),
      exp(angle_beta[5]),
      exp(angle_beta[6])
    )
  )
}

calc_densities <- function(param_table, max_step = 6000, n = 500){
  step_x <- seq(0, max_step, length.out = n)
  angle_x <- seq(-pi, pi, length.out = n)
  
  step_dens <- purrr::map_dfr(seq_len(nrow(param_table)), function(i){
    p <- param_table[i, ]
    shape <- p$step_mean^2 / p$step_sd^2
    scale <- p$step_sd^2 / p$step_mean
    
    data.frame(
      x = step_x,
      density = dgamma(step_x, shape = shape, scale = scale),
      state = p$state
    )
  })
  
  angle_dens <- purrr::map_dfr(seq_len(nrow(param_table)), function(i){
    p <- param_table[i, ]
    d_angle <- if(p$angle_conc < 1e-8){
      rep(1 / (2 * pi), length(angle_x))
    } else {
      exp(p$angle_conc * cos(angle_x)) / (2 * pi * besselI(p$angle_conc, 0))
    }
    
    data.frame(
      x = angle_x,
      density = d_angle,
      state = p$state
    )
  })
  
  list(step = step_dens, angle = angle_dens)
}

cols <- HMM_STATE_COLOURS
state_labels <- HMM_STATE_LABELS

param_table <- get_params(best)
dens <- calc_densities(param_table, max_step = 6100)

draw_order <- c("water", "ARS", "undirected", "directed")

dens_step <- dens$step %>%
  dplyr::mutate(state = factor(state, levels = draw_order)) %>%
  dplyr::arrange(state)

dens_angle <- dens$angle %>%
  dplyr::mutate(state = factor(state, levels = draw_order)) %>%
  dplyr::arrange(state)

p_step <- ggplot() +
  geom_histogram(data = bg_step, aes(x = step, y = after_stat(density)), bins = 70, fill = "grey80", color = NA) +
  geom_line(data = dens_step, aes(x = x, y = density, colour = state), linewidth = 1) +
  scale_colour_manual(values = cols, breaks = stateNames, labels = state_labels) +
  coord_cartesian(xlim = c(0, 6100)) +
  labs(x = "Step length (m)", y = "Density", colour = NULL) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 20, margin = margin(t = 10)),
    axis.title.y = element_text(size = 20, margin = margin(r = 12)),
    axis.text = element_text(size = 18),
    legend.title = element_blank(),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    plot.margin = margin(t = 5, r = 3, b = 5, l = 5)
  )

p_angle <- ggplot() +
  geom_histogram(data = bg_angle, aes(x = angle, y = after_stat(density)), bins = 35, fill = "grey80", color = NA) +
  geom_line(data = dens_angle, aes(x = x, y = density, colour = state), linewidth = 1) +
  scale_colour_manual(values = cols, breaks = stateNames, labels = state_labels) +
  scale_x_continuous(breaks = c(-pi, -pi / 2, 0, pi / 2, pi), labels = c(expression(-pi), expression(-pi/2), "0", expression(pi/2), expression(pi))) +
  labs(x = "Turning angle (radians)", y = "", colour = NULL) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 20, margin = margin(t = 10)),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    axis.ticks.y = element_line(),
    axis.line = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 3)
  )

legend_dist <- cowplot::get_legend(
  p_angle +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.title = element_blank(),
      legend.text = element_text(size = 18)
    ) +
    guides(
      colour = guide_legend(title = NULL, nrow = 1)
    )
)

p_step_noleg <- p_step + theme(legend.position = "none")
p_angle_noleg <- p_angle + theme(legend.position = "none")

p_main <- cowplot::ggdraw() +
  cowplot::draw_plot(
    p_step_noleg,
    x = 0.00,
    y = 0,
    width = 0.5,
    height = 1
  ) +
  cowplot::draw_plot(
    p_angle_noleg,
    x = 0.52,
    y = 0,
    width = 0.43,
    height = 1
  )

legend_left <- cowplot::ggdraw() +
  cowplot::draw_plot(
    legend_dist,
    x = -0.1,
    y = 0,
    width = 0.75,
    height = 1
  )

p_dist <- cowplot::plot_grid(
  p_main,
  legend_left,
  ncol = 1,
  rel_heights = c(1, 0.14)
)

print(p_dist)

save_pdf(
  filename = here::here("figure", "HMM_distribution.pdf"),
  plot = p_dist,
  size = FIGURE_SIZES$hmm_distribution
)
```



### Figure 2: (a)step, (b)angle

plot

```{r}
plot_point_df <- par_by_id %>%
  dplyr::mutate(id = as.character(id)) %>%
  dplyr::select(
    -dplyr::any_of(c(
      "success_rate",
      "success_rate.x",
      "success_rate.y"
    ))
  ) %>%
  dplyr::left_join(success_rate_by_id, by = "id") %>%
  dplyr::select(
    id,
    success_rate,
    step_mean_directed,
    step_mean_undirected,
    angle_conc_directed,
    angle_conc_undirected
  ) %>%
  tidyr::pivot_longer(
    cols = -c(id, success_rate),
    names_to = c("parameter_raw", "state"),
    names_pattern = "(step_mean|angle_conc)_(.*)",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    state_lab = dplyr::case_when(
      state == "directed" ~ "Directed flight",
      state == "undirected" ~ "Undirected flight"
    ),
    parameter_lab = dplyr::case_when(
      parameter_raw == "step_mean" ~ "Mean step length (m)",
      parameter_raw == "angle_conc" ~ "Turning-angle concentration"
    ),
    state_lab = factor(
      state_lab,
      levels = c("Directed flight", "Undirected flight")
    ),
    parameter_lab = factor(
      parameter_lab,
      levels = c(
        "Mean step length (m)",
        "Turning-angle concentration"
      )
    )
  ) %>%
  dplyr::filter(
    !is.na(success_rate),
    !is.na(value),
    !is.na(state_lab),
    !is.na(parameter_lab)
  )
```


HMM parameters

```{r}
success_rate_original <- score$success_rate
score_mean <- mean(success_rate_original, na.rm = TRUE)
score_sd   <- sd(success_rate_original, na.rm = TRUE)

score_sc_range <- range(best$data$score_sc, na.rm = TRUE)
score_seq <- seq(score_sc_range[1], score_sc_range[2], length.out = 200)

step_beta  <- unlist(best$mle$step)
angle_beta <- unlist(best$mle$angle)
Sigma <- best$mod$Sigma
n_step <- length(step_beta)

state_levels <- c("directed", "undirected")

cols_2 <- HMM_STATE_COLOURS[c("directed", "undirected")]

# Function to calculate fitted values and 95% CI

make_ci_param <- function(b0, b1, V, x, state, parameter) {
  
  eta <- b0 + b1 * x
  se_eta <- sqrt(V[1, 1] + x^2 * V[2, 2] + 2 * x * V[1, 2])
  z <- qnorm(0.975)
  
  tibble::tibble(
    score_sc = x,
    success_rate = x * score_sd + score_mean,
    state = state,
    parameter = parameter,
    fit = exp(eta),
    lower = exp(eta - z * se_eta),
    upper = exp(eta + z * se_eta)
  )
}


# Build prediction table from covariate HMM

param_df <- dplyr::bind_rows(
  make_ci_param(
    b0 = step_beta[1],
    b1 = step_beta[2],
    V = Sigma[c(1, 2), c(1, 2)],
    x = score_seq,
    state = "directed",
    parameter = "step_mean"
  ),
  make_ci_param(
    b0 = step_beta[3],
    b1 = step_beta[4],
    V = Sigma[c(3, 4), c(3, 4)],
    x = score_seq,
    state = "undirected",
    parameter = "step_mean"
  ),
  make_ci_param(
    b0 = angle_beta[1],
    b1 = angle_beta[2],
    V = Sigma[n_step + c(1, 2), n_step + c(1, 2)],
    x = score_seq,
    state = "directed",
    parameter = "angle_concentration"
  ),
  make_ci_param(
    b0 = angle_beta[3],
    b1 = angle_beta[4],
    V = Sigma[n_step + c(3, 4), n_step + c(3, 4)],
    x = score_seq,
    state = "undirected",
    parameter = "angle_concentration"
  )
) %>%
  dplyr::mutate(
    state = factor(state, levels = state_levels),
    parameter = factor(
      parameter,
      levels = c("step_mean", "angle_concentration")
    )
  )


# Plot function

if (!requireNamespace("ggh4x", quietly = TRUE)) {
  stop("Package 'ggh4x' is required. Install it before rendering this notebook.")
}

plot_param_df <- param_df %>%
  dplyr::filter(
    state %in% c("directed", "undirected"),
    parameter %in% c("step_mean", "angle_concentration")
  ) %>%
  dplyr::mutate(
    state_lab = dplyr::case_when(
      state == "directed" ~ "Directed flight",
      state == "undirected" ~ "Undirected flight"
    ),
    parameter_lab = dplyr::case_when(
      parameter == "step_mean" ~ "Mean step length (m)",
      parameter == "angle_concentration" ~ "Turning-angle concentration"
    ),
    state_lab = factor(
      state_lab,
      levels = c("Directed flight", "Undirected flight")
    ),
    parameter_lab = factor(
      parameter_lab,
      levels = c("Mean step length (m)", "Turning-angle concentration")
    )
  )

cols_2_lab <- HMM_LABEL_COLOURS[c("Directed flight", "Undirected flight")]

p_hmm_score_effects_2x2 <- ggplot(
  plot_param_df,
  aes(x = success_rate, y = fit, colour = state_lab, fill = state_lab)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.20,
    colour = NA
  ) +
  geom_line(
    linewidth = 1.3
  ) +
  geom_point(
  data = plot_point_df,
  aes(
    x = success_rate,
    y = value,
    colour = state_lab
  ),
  inherit.aes = FALSE,
  shape = 21,
  fill = "white",
  stroke = 0.8,
  size = 3,
  alpha = 0.9
) +
  ggh4x::facet_grid2(
    rows = vars(parameter_lab),
    cols = vars(state_lab),
    scales = "free_y",
    independent = "y",
    axes = "x",
    remove_labels = "none",
    switch = "y"
  ) +
  ggh4x::facetted_pos_scales(
    y = list(
      parameter_lab == "Mean step length (m)" ~
        scale_y_continuous(
          labels = scales::label_number(
            accuracy = 1,
            big.mark = ""
          )
        ),
      parameter_lab == "Turning-angle concentration" ~
        scale_y_continuous(
          labels = scales::label_number(
            accuracy = 0.1
          )
        )
    )
  ) +
  scale_colour_manual(values = cols_2_lab) +
  scale_fill_manual(values = cols_2_lab) +
  scale_x_continuous(
    breaks = c(0, 0.2, 0.4, 0.6),
    labels = scales::label_number(accuracy = 0.1)
  ) +
  labs(
    x = "Cylinder-task success rate",
    y = NULL
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(
      size = 24,
      margin = margin(t = 0, b = 15)
    ),
    strip.text.y.left = element_text(
      size = 20,
      angle = 90,
      margin = margin(r = 10)
    ),
    
    axis.title.x = element_text(
      size = 20,
      margin = margin(t = 12)
    ),
    axis.text = element_text(size = 16),
    axis.line = element_blank(),
    axis.ticks.x = element_line(),
    axis.text.x = element_text(size = 16),
    
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.8
    ),
    
    panel.spacing.x = grid::unit(1.0, "lines"),
    panel.spacing.y = grid::unit(1.5, "lines"),
    
    plot.margin = margin(t = 6, r = 8, b = 6, l = 6)
  )

p_hmm_score_effects_2x2

save_pdf(
  filename = here::here("figure", "HMM_score_effects.pdf"),
  plot = p_hmm_score_effects_2x2,
  size = FIGURE_SIZES$hmm_score_effects
)
```


## Figure 3: Behavioural residency time and stationary probability

Behavioural residency time data frame
```{r}
# Setting 

dt_min <- 5

state_breaks <- c("directed", "undirected", "ARS", "water")

state_labels <- HMM_STATE_LABELS_ON_WATER
state_cols <- HMM_STATE_COLOURS

score_mean <- mean(score$success_rate, na.rm = TRUE)
score_sd   <- sd(score$success_rate, na.rm = TRUE)

score_seq <- seq(
  min(best$data$score_sc, na.rm = TRUE),
  max(best$data$score_sc, na.rm = TRUE),
  length.out = 100
)

score_grid <- tibble::tibble(
  score_sc = score_seq,
  success_rate = score_sc * score_sd + score_mean
)
```


```{r}
dwell_from_p <- function(p, dt_min = 5) {
  dplyr::if_else(
    is.finite(p) & p >= 0 & p < 1,
    dt_min / (1 - p),
    NA_real_
  )
}

extract_self_transition_ci <- function(model, cov_row, stateNames, alpha = 0.95) {
  
  ci <- momentuHMM::CIreal(
    model,
    covs = cov_row,
    parms = "gamma",
    alpha = alpha
  )
  
  gamma_ci <- ci$gamma
  
  est_mat   <- gamma_ci$est
  lower_mat <- gamma_ci$lower
  upper_mat <- gamma_ci$upper
  
  rownames(est_mat) <- colnames(est_mat) <- stateNames
  rownames(lower_mat) <- colnames(lower_mat) <- stateNames
  rownames(upper_mat) <- colnames(upper_mat) <- stateNames
  
  tibble::tibble(
    state = stateNames,
    self_p = diag(est_mat),
    self_p_lower = diag(lower_mat),
    self_p_upper = diag(upper_mat)
  )
}

dwell_df <- purrr::map_dfr(seq_len(nrow(score_grid)), function(i) {
  
  cov_row <- score_grid[i, "score_sc", drop = FALSE]
  
  extract_self_transition_ci(
    model = best,
    cov_row = cov_row,
    stateNames = stateNames,
    alpha = 0.95
  ) %>%
    dplyr::mutate(
      score_sc = score_grid$score_sc[i],
      success_rate = score_grid$success_rate[i]
    )
}) %>%
  dplyr::mutate(
    state = factor(state, levels = state_breaks),
    
    fit = dwell_from_p(self_p, dt_min),
    lower_raw = dwell_from_p(self_p_lower, dt_min),
    upper_raw = dwell_from_p(self_p_upper, dt_min),
    
    lower = pmin(lower_raw, upper_raw, na.rm = FALSE),
    upper = pmax(lower_raw, upper_raw, na.rm = FALSE),
    
    has_ci = is.finite(lower) & is.finite(upper)
  )
```

Stationary probability data frame

```{r}
# Clay et al. 2020 (ci.stationary)

#######
ci.stationary <- function(model, cov, alpha = 0.95) {
  gridLength <- nrow(cov)
  model2 <- momentuHMM:::delta_bc(model)
  nbStates <- length(model2$stateNames)
  Sigma <- model2$mod$Sigma

  lci <- matrix(NA, gridLength, nbStates)
  uci <- matrix(NA, gridLength, nbStates)

  formula <- model2$conditions$formula
  newForm <- momentuHMM:::newFormulas(formula, nbStates)
  newformula <- newForm$newformula

  nbCovs <- ncol(model.matrix(newformula, model2$data)) - 1

  gamInd <- (length(model2$mod$estimate) - (nbCovs + 1) * nbStates * (nbStates - 1) + 1):length(model2$mod$estimate) -
    ncol(model2$covsDelta) * (nbStates - 1) * (!model2$conditions$stationary)

  rawCovs <- model2$rawCovs
  rawCovs <- rawCovs[, order(names(rawCovs)), drop = FALSE]
  tempCovs <- cov
  tempCovs <- tempCovs[, sort(names(rawCovs)), drop = FALSE]

  for (i in which(unlist(lapply(rawCovs, is.factor)))) {
    tempCovs[[i]] <- factor(tempCovs[[i]], levels = levels(rawCovs[, i]))
  }

  tmpSplineInputs <- momentuHMM:::getSplineFormula(
    newformula,
    model2$data,
    tempCovs
  )

  desMat <- model.matrix(
    tmpSplineInputs$formula,
    data = tmpSplineInputs$covs
  )

  probs <- as.data.frame(
    momentuHMM::stationary(model2, covs = desMat)
  )

  for (state in seq_len(nbStates)) {
    dN <- t(apply(desMat, 1, function(x) {
      numDeriv::grad(
        momentuHMM:::get_stat,
        model2$mod$estimate[gamInd][unique(c(model2$conditions$betaCons))],
        covs = matrix(x, nrow = 1),
        nbStates = nbStates,
        i = state,
        betaRef = model2$conditions$betaRef,
        betaCons = model2$conditions$betaCons,
        workBounds = model2$conditions$workBounds$beta
      )
    }))

    se <- t(apply(dN, 1, function(x) {
      suppressWarnings(
        sqrt(
          x %*%
            Sigma[
              gamInd[unique(c(model2$conditions$betaCons))],
              gamInd[unique(c(model2$conditions$betaCons))]
            ] %*%
            x
        )
      )
    }))

    z <- stats::qnorm(1 - (1 - alpha) / 2)

    lci[, state] <- 1 / (
      1 + exp(-(
        log(probs[, state] / (1 - probs[, state])) -
          z * (1 / (probs[, state] - probs[, state]^2)) * se
      ))
    )

    uci[, state] <- 1 / (
      1 + exp(-(
        log(probs[, state] / (1 - probs[, state])) +
          z * (1 / (probs[, state] - probs[, state]^2)) * se
      ))
    )
  }

  lci <- as.data.frame(lci)
  names(lci) <- model$stateNames

  uci <- as.data.frame(uci)
  names(uci) <- model$stateNames

  list(
    lower = lci,
    upper = uci
  )
}
######

stationary_est <- as.data.frame(
  momentuHMM::stationary(
    model = best,
    covs = score_grid %>% dplyr::select(score_sc)
  )
)

stationary_ci <- ci.stationary(
  model = best,
  cov = score_grid %>% dplyr::select(score_sc),
  alpha = 0.95
)

stationary_df <- tibble::tibble(
  score_sc = score_grid$score_sc,
  success_rate = score_grid$success_rate
) %>%
  dplyr::bind_cols(stationary_est) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(stateNames),
    names_to = "state",
    values_to = "fit"
  ) %>%
  dplyr::left_join(
    tibble::tibble(
      score_sc = score_grid$score_sc,
      success_rate = score_grid$success_rate
    ) %>%
      dplyr::bind_cols(stationary_ci$lower) %>%
      tidyr::pivot_longer(
        cols = dplyr::all_of(stateNames),
        names_to = "state",
        values_to = "lower"
      ),
    by = c("score_sc", "success_rate", "state")
  ) %>%
  dplyr::left_join(
    tibble::tibble(
      score_sc = score_grid$score_sc,
      success_rate = score_grid$success_rate
    ) %>%
      dplyr::bind_cols(stationary_ci$upper) %>%
      tidyr::pivot_longer(
        cols = dplyr::all_of(stateNames),
        names_to = "state",
        values_to = "upper"
      ),
    by = c("score_sc", "success_rate", "state")
  ) %>%
  dplyr::mutate(
    state = factor(state, levels = state_breaks),
    has_ci = is.finite(lower) & is.finite(upper)
  )
```


```{r}
x_max <- max(score$success_rate, na.rm = TRUE)

# ------------------------------------------------------------
# Theme
# ------------------------------------------------------------

theme_time_state <- theme_classic() +
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 22, margin = margin(t = 12)),
    axis.title.y = element_text(size = 22, margin = margin(r = 12)),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 20),
    axis.line = element_blank(),
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.8
    )
  )

# ------------------------------------------------------------
# Plot
# ------------------------------------------------------------

plot_time_state <- function(df, ylab_text, panel_label) {
  
  ggplot(df, aes(x = success_rate, y = fit)) +
    geom_ribbon(
      data = df %>% dplyr::filter(has_ci),
      aes(
        ymin = lower,
        ymax = upper,
        fill = state
      ),
      alpha = 0.20,
      colour = NA
    ) +
    geom_line(
      aes(colour = state),
      linewidth = 1.3
    ) +
    annotate(
      "text",
      x = -Inf,
      y = Inf,
      label = panel_label,
      parse = TRUE,
      hjust = -0.3,
      vjust = 1.6,
      size = 9
    ) +
    scale_x_continuous(
      limits = c(0, x_max),
      breaks = c(0, 0.2, 0.4, 0.6),
      labels = scales::label_number(accuracy = 0.1)
    ) +
    scale_colour_manual(
      name = "State",
      values = state_cols,
      breaks = state_breaks,
      labels = state_labels
    ) +
    scale_fill_manual(
      name = "State",
      values = state_cols,
      breaks = state_breaks,
      labels = state_labels
    ) +
    labs(
      x = "Cylinder-task success rate",
      y = ylab_text,
      colour = "State",
      fill = "State"
    ) +
    theme_time_state
}

p_dwell <- plot_time_state(
  df = dwell_df,
  ylab_text = "Expected state duration (min)",
  panel_label = "'(' * italic(a) * ')'"
)

p_stationary <- plot_time_state(
  df = stationary_df,
  ylab_text = "Stationary probability",
  panel_label = "'(' * italic(b) * ')'"
)

# ------------------------------------------------------------
# Combine and save
# ------------------------------------------------------------

legend_bottom <- cowplot::get_legend(
  p_stationary +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.title = element_blank()
    ) +
    guides(
      colour = guide_legend(nrow = 1),
      fill = guide_legend(nrow = 1)
    )
)

p_main <- cowplot::plot_grid(
  p_dwell + theme(
    legend.position = "none",
    plot.margin = margin(t = 5, r = 14, b = 5, l = 5)
  ),
  p_stationary + theme(
    legend.position = "none",
    plot.margin = margin(t = 5, r = 5, b = 5, l = 14)
  ),
  nrow = 1,
  rel_widths = c(1, 1),
  align = "hv",
  axis = "tb"
)

legend_left <- cowplot::plot_grid(
  NULL,
  legend_bottom,
  NULL,
  nrow = 1,
  rel_widths = c(0.08, 0.50, 0.42)
)

p_time_stationary <- cowplot::plot_grid(
  p_main,
  legend_left,
  ncol = 1,
  rel_heights = c(1, 0.13)
)

p_time_stationary

save_pdf(
  filename = here::here("figure", "HMM_dwell_time_stationary_probability.pdf"),
  plot = p_time_stationary,
  size = FIGURE_SIZES$hmm_residency
)
```


### Figure S2

Prepare individual-level HMM parameter data

```{r}
required_hmm_objects <- c("model_by_id", "id_model_summary", "state_prop_by_id", "par_by_id")
missing_hmm_objects <- required_hmm_objects[!vapply(required_hmm_objects, exists, logical(1))]
if (length(missing_hmm_objects) > 0) {
  stop("Loaded RDS is missing required individual HMM objects: ", paste(missing_hmm_objects, collapse = ", "))
}
id_model_summary
state_prop_by_id
par_by_id
```
Fit GLMs and create predictions

```{r}
out_dir <- here::here("figure", "individual_HMM")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ------------------------------------------------------------
# Settings
# ------------------------------------------------------------

state_levels <- c("directed", "undirected")

cols_2 <- HMM_STATE_COLOURS[c("directed", "undirected")]

# ------------------------------------------------------------
# Prepare individual-level HMM parameter data
# ------------------------------------------------------------

ind_param_df <- par_by_id %>%
  dplyr::mutate(id = as.character(id)) %>%
  dplyr::select(
    -dplyr::any_of(c(
      "success_rate",
      "success_rate.x",
      "success_rate.y"
    ))
  ) %>%
  dplyr::left_join(success_rate_by_id, by = "id") %>%
  dplyr::select(
    id,
    success_rate,
    step_mean_directed,
    step_mean_undirected,
    angle_conc_directed,
    angle_conc_undirected
  ) %>%
  tidyr::pivot_longer(
    cols = -c(id, success_rate),
    names_to = c("parameter_raw", "state"),
    names_pattern = "(step_mean|angle_conc)_(.*)",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    state = factor(state, levels = state_levels),
    parameter = dplyr::recode(
      parameter_raw,
      step_mean = "step_mean",
      angle_conc = "angle_concentration"
    ),
    parameter = factor(
      parameter,
      levels = c("step_mean", "angle_concentration")
    )
  ) %>%
  dplyr::filter(
    !is.na(success_rate),
    !is.na(value),
    !is.na(state)
  )

# ------------------------------------------------------------
# Fit GLMs and create prediction table
# ------------------------------------------------------------

make_glm_pred <- function(dat) {
  
  dat <- dat %>%
    dplyr::filter(value > 0)
  
  m <- glmmTMB::glmmTMB(
    value ~ success_rate,
    data = dat,
    family = lognormal(link = "log")
  )
  
  newdat <- tibble::tibble(
    success_rate = seq(
      min(dat$success_rate, na.rm = TRUE),
      max(dat$success_rate, na.rm = TRUE),
      length.out = 100
    )
  )
  
  pr <- predict(
    m,
    newdata = newdat,
    se.fit = TRUE,
    type = "link"
  )
  
  newdat %>%
    dplyr::mutate(
      fit = exp(pr$fit),
      lower = exp(pr$fit - qnorm(0.975) * pr$se.fit),
      upper = exp(pr$fit + qnorm(0.975) * pr$se.fit)
    )
}

glm_results <- ind_param_df %>%
  dplyr::filter(value > 0) %>%
  dplyr::group_by(parameter, state) %>%
  tidyr::nest() %>%
  dplyr::mutate(
    model = purrr::map(
      data,
      ~ glmmTMB::glmmTMB(
        value ~ success_rate,
        data = .x,
        family = lognormal(link = "log")
      )
    ),
    pred = purrr::map(data, make_glm_pred),
    model_summary = purrr::map(model, summary)
  ) %>%
  dplyr::ungroup()

glm_pred_df <- glm_results %>%
  dplyr::select(parameter, state, pred) %>%
  tidyr::unnest(pred)

glm_summary_list <- glm_results %>%
  dplyr::select(parameter, state, model_summary)

# ------------------------------------------------------------
# Print model summaries
# ------------------------------------------------------------

glm_summary_list %>%
  dplyr::mutate(
    label = paste(parameter, state, sep = " | ")
  ) %>%
  dplyr::select(label, model_summary) %>%
  purrr::pwalk(function(label, model_summary) {
    
    cat("\n\n")
    cat("============================================================\n")
    cat(label, "\n")
    cat("============================================================\n\n")
    
    print(model_summary)
  })
```


Plot

```{r}
# ------------------------------------------------------------
# Individual-level HMM GLM figure: compact facet version
# ------------------------------------------------------------

if (!requireNamespace("ggh4x", quietly = TRUE)) {
  stop("Package 'ggh4x' is required. Install it before rendering this notebook.")
}

plot_pred_df <- glm_pred_df %>%
  dplyr::filter(
    state %in% c("directed", "undirected"),
    parameter %in% c("step_mean", "angle_concentration")
  ) %>%
  dplyr::mutate(
    state_lab = dplyr::case_when(
      state == "directed" ~ "Directed flight",
      state == "undirected" ~ "Undirected flight"
    ),
    parameter_lab = dplyr::case_when(
      parameter == "step_mean" ~ "Step length mean (m)",
      parameter == "angle_concentration" ~ "Angle concentration (\u03BA)"
    ),
    state_lab = factor(
      state_lab,
      levels = c("Directed flight", "Undirected flight")
    ),
    parameter_lab = factor(
      parameter_lab,
      levels = c("Step length mean (m)", "Angle concentration (\u03BA)")
    )
  )

plot_point_df <- ind_param_df %>%
  dplyr::filter(
    state %in% c("directed", "undirected"),
    parameter %in% c("step_mean", "angle_concentration")
  ) %>%
  dplyr::mutate(
    state_lab = dplyr::case_when(
      state == "directed" ~ "Directed flight",
      state == "undirected" ~ "Undirected flight"
    ),
    parameter_lab = dplyr::case_when(
      parameter == "step_mean" ~ "Step length mean (m)",
      parameter == "angle_concentration" ~ "Angle concentration (\u03BA)"
    ),
    state_lab = factor(
      state_lab,
      levels = c("Directed flight", "Undirected flight")
    ),
    parameter_lab = factor(
      parameter_lab,
      levels = c("Step length mean (m)", "Angle concentration (\u03BA)")
    )
  )

cols_2_lab <- HMM_LABEL_COLOURS[c("Directed flight", "Undirected flight")]

p_ind_glm_2x2 <- ggplot(
  plot_pred_df,
  aes(x = success_rate, y = fit, colour = state_lab, fill = state_lab)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.2,
    colour = NA
  ) +
  geom_line(linewidth = 1.3) +
  geom_point(
    data = plot_point_df,
    aes(x = success_rate, y = value, colour = state_lab),
    inherit.aes = FALSE,
    size = 3,
    shape = 21,
    fill = "white",
    stroke = 0.8,
    alpha = 0.9
  ) +
  ggh4x::facet_grid2(
  rows = vars(parameter_lab),
  cols = vars(state_lab),
  scales = "free_y",
  independent = "y",
  axes = "x",
  remove_labels = "none",
  switch = "y"
) +
  ggh4x::facetted_pos_scales(
  y = list(
    parameter_lab == "Step length mean (m)" ~
      scale_y_continuous(
        labels = scales::label_number(
          accuracy = 1,
          big.mark = ""
        )
      ),
    parameter_lab == "Angle concentration (\u03BA)" ~
      scale_y_continuous(
        labels = scales::label_number(
          accuracy = 0.1
        )
      )
  )
  ) +
  scale_colour_manual(values = cols_2_lab) +
  scale_fill_manual(values = cols_2_lab) +
  scale_x_continuous(
    breaks = c(0, 0.2, 0.4, 0.6),
    labels = scales::label_number(accuracy = 0.1)
  ) +
  labs(
    x = "Success rate",
    y = NULL
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(size = 24,margin = margin(t = 0, b = 15)),
    strip.text.y.left = element_text(size = 20, angle = 90, margin = margin(r = 10)),
    axis.title.x = element_text(size = 20, margin = margin(t = 12)),
    axis.text = element_text(size = 16),
    axis.line = element_blank(),
    axis.ticks.x = element_line(),
    axis.text.x = element_text(size = 16),
    
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.8
    ),
    
    panel.spacing.x = grid::unit(1.0, "lines"),
    panel.spacing.y = grid::unit(1.5, "lines"),
    
    plot.margin = margin(t = 6, r = 8, b = 6, l = 6)
  )

p_ind_glm_2x2

save_pdf(
  filename = here::here("figure", "individual_HMM_GLM.pdf"),
  plot = p_ind_glm_2x2,
  size = FIGURE_SIZES$individual_hmm_glm
)
```


# Additional stationary-distribution output
# Figure: proportion of Viterbi states by individual
Covariate HMM
```{r}
# Viterbi states from the covariate HMM
vit <- momentuHMM::viterbi(best)

if (length(vit) != nrow(Data_HMM)) {
  stop("Length of Viterbi states does not match nrow(Data_HMM).")
}

viterbi_prop_df <- Data_HMM %>%
  dplyr::mutate(
    id = as.character(id),
    viterbi_state = if (is.numeric(vit) || is.integer(vit)) {
      factor(vit, levels = seq_along(stateNames), labels = stateNames)
    } else {
      factor(vit, levels = stateNames)
    }
  ) %>%
  dplyr::filter(!is.na(viterbi_state)) %>%
  dplyr::count(id, state = viterbi_state, name = "n") %>%
  tidyr::complete(
    id,
    state = factor(stateNames, levels = stateNames),
    fill = list(n = 0)
  ) %>%
  dplyr::group_by(id) %>%
  dplyr::mutate(
    prop = n / sum(n)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(success_rate_by_id, by = "id") %>%
  dplyr::filter(!is.na(success_rate)) %>%
  dplyr::mutate(
    state_lab = dplyr::recode(
      as.character(state),
      directed = "Directed flight",
      undirected = "Undirected flight",
      ARS = "ARS",
      water = "Water"
    ),
    # Stack order:
    # ggplot stacks from bottom to top by factor order.
    # To make the visible order from top to bottom:
    # Directed -> Undirected -> ARS -> Water,
    # set bottom-to-top order as:
    # Water -> ARS -> Undirected -> Directed.
    state_lab = factor(
      state_lab,
      levels = c(
        "Water",
        "ARS",
        "Undirected flight",
        "Directed flight"
      )
    )
  )

# Order individuals by increasing success rate
id_success_order <- success_rate_by_id %>%
  dplyr::filter(id %in% viterbi_prop_df$id) %>%
  dplyr::arrange(success_rate, id) %>%
  dplyr::mutate(
    id_success_lab = scales::number(
      success_rate,
      accuracy = 0.01
    )
  ) %>%
  dplyr::select(id, success_rate, id_success_lab)

viterbi_prop_df <- viterbi_prop_df %>%
  dplyr::mutate(
    id_success_fct = factor(
      id,
      levels = id_success_order$id
    )
  )

# Colours consistent with earlier figures
cols_4_lab <- HMM_LABEL_COLOURS

p_viterbi_state_prop_by_success <- ggplot(
  viterbi_prop_df,
  aes(x = id_success_fct, y = prop, fill = state_lab)
) +
  geom_col(width = 0.55, alpha = 0.5) +
  scale_fill_manual(
    values = cols_4_lab,
    breaks = c("Directed flight", "Undirected flight", "ARS", "Water"),
    guide = guide_legend(nrow = 1, byrow = TRUE)
  ) +
  scale_x_discrete(
    labels = stats::setNames(
      id_success_order$id_success_lab,
      id_success_order$id
    )
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = expression(italic("(a)") ~ "Covariate HMM"),
    x = "Success rate",
    y = "Proportion of Viterbi states",
    fill = NULL
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 17, hjust = 0),
    axis.title.x = element_text(size = 18, margin = margin(t = 12)),
    axis.title.y = element_text(size = 18, margin = margin(r = 12)),
    axis.text.x = element_text(size = 11, angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size = 13),
    axis.line = element_blank(),
    axis.ticks = element_line(linewidth = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    plot.margin = margin(t = 6, r = 8, b = 6, l = 6)
  )
```

## Results from individual-level HMMs
```{r}
# ------------------------------------------------------------
# Figure: proportion of Viterbi states by individual
# Individual-level HMMs
# ------------------------------------------------------------

# ------------------------------------------------------------
# Extract Viterbi states from each individual-level HMM
# ------------------------------------------------------------

viterbi_prop_ind_df <- purrr::imap_dfr(model_by_id, function(m_i, id_i) {
  
  if (is.null(m_i)) {
    return(NULL)
  }
  
  dat_i <- m_i$data
  vit_i <- momentuHMM::viterbi(m_i)
  
  if (length(vit_i) != nrow(dat_i)) {
    stop(paste0(
      "Length of Viterbi states does not match nrow(data) for id: ",
      id_i
    ))
  }
  
  dat_i %>%
    dplyr::mutate(
      id = as.character(id_i),
      viterbi_state = if (is.numeric(vit_i) || is.integer(vit_i)) {
        factor(
          vit_i,
          levels = seq_along(stateNames_ind),
          labels = stateNames_ind
        )
      } else {
        factor(
          vit_i,
          levels = stateNames_ind
        )
      }
    ) %>%
    dplyr::select(id, viterbi_state)
})

# ------------------------------------------------------------
# Calculate state proportions by individual
# ------------------------------------------------------------

viterbi_prop_ind_df <- viterbi_prop_ind_df %>%
  dplyr::filter(!is.na(viterbi_state)) %>%
  dplyr::count(id, state = viterbi_state, name = "n") %>%
  tidyr::complete(
    id,
    state = factor(stateNames_ind, levels = stateNames_ind),
    fill = list(n = 0)
  ) %>%
  dplyr::group_by(id) %>%
  dplyr::mutate(
    prop = n / sum(n)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(success_rate_by_id, by = "id") %>%
  dplyr::filter(!is.na(success_rate)) %>%
  dplyr::mutate(
    state_lab = dplyr::recode(
      as.character(state),
      directed = "Directed flight",
      undirected = "Undirected flight",
      ARS = "ARS",
      water = "Water"
    ),
    # Stack order:
    # ggplot stacks from bottom to top by factor order.
    # Visible order from top to bottom:
    # Directed -> Undirected -> ARS -> Water
    state_lab = factor(
      state_lab,
      levels = c(
        "Water",
        "ARS",
        "Undirected flight",
        "Directed flight"
      )
    )
  )

# ------------------------------------------------------------
# Order individuals by increasing success rate
# ------------------------------------------------------------

id_success_order_ind <- success_rate_by_id %>%
  dplyr::filter(id %in% viterbi_prop_ind_df$id) %>%
  dplyr::arrange(success_rate, id) %>%
  dplyr::mutate(
    id_success_lab = scales::number(
      success_rate,
      accuracy = 0.01
    )
  ) %>%
  dplyr::select(id, success_rate, id_success_lab)

viterbi_prop_ind_df <- viterbi_prop_ind_df %>%
  dplyr::mutate(
    id_success_fct = factor(
      id,
      levels = id_success_order_ind$id
    )
  )

# Colours consistent with earlier figures
cols_4_lab <- HMM_LABEL_COLOURS

# ------------------------------------------------------------
# Plot
# ------------------------------------------------------------

p_viterbi_state_prop_by_success_ind <- ggplot(
  viterbi_prop_ind_df,
  aes(x = id_success_fct, y = prop, fill = state_lab)
) +
  geom_col(width = 0.55, alpha = 0.5) +
  scale_fill_manual(
    values = cols_4_lab,
    breaks = c("Directed flight", "Undirected flight", "ARS", "Water"),
    guide = guide_legend(nrow = 1, byrow = TRUE)
  ) +
  scale_x_discrete(
    labels = stats::setNames(
      id_success_order_ind$id_success_lab,
      id_success_order_ind$id
    )
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = expression(italic("(b)") ~ "Individually fitted HMMs"),
    x = "Success rate",
    y = "Proportion of Viterbi states",
    fill = NULL
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 17, hjust = 0),
    axis.title.x = element_text(size = 18, margin = margin(t = 12)),
    axis.title.y = element_text(size = 18, margin = margin(r = 12)),
    axis.text.x = element_text(size = 11, angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size = 13),
    axis.line = element_blank(),
    axis.ticks = element_line(linewidth = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    plot.margin = margin(t = 6, r = 8, b = 6, l = 6)
  )
```

```{r}
p_cov_panel <- p_viterbi_state_prop_by_success +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(
      size = 22,
      hjust = -0.16,
      vjust = 1,
      margin = ggplot2::margin(t = 0, b = 8)
    ),
    axis.text.x = ggplot2::element_text(
      size = 12,
      angle = 45,
      hjust = 1,
      vjust = 1,
      margin = ggplot2::margin(t = 0)
    ),
    axis.text.y = ggplot2::element_text(
      size = 17,
      margin = ggplot2::margin(r = 0)
    ),
    axis.title.x = ggplot2::element_text(
      size = 18,
      margin = ggplot2::margin(t = 10)
    ),
    axis.title.y = ggplot2::element_text(
      size = 18,
      margin = ggplot2::margin(r = 2)
    ),
    plot.margin = ggplot2::margin(t = 14, r = 8, b = 6, l = 34)
  )

p_ind_panel <- p_viterbi_state_prop_by_success_ind +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(
      size = 22,
      hjust = -0.16,
      vjust = 1,
      margin = ggplot2::margin(t = 0, b = 8)
    ),
    axis.text.x = ggplot2::element_text(
      size = 12,
      angle = 45,
      hjust = 1,
      vjust = 1,
      margin = ggplot2::margin(t = 0)
    ),
    axis.text.y = ggplot2::element_text(
      size = 17,
      margin = ggplot2::margin(r = 0)
    ),
    axis.title.x = ggplot2::element_text(
      size = 18,
      margin = ggplot2::margin(t = 10)
    ),
    axis.title.y = ggplot2::element_text(
      size = 18,
      margin = ggplot2::margin(r = 2)
    ),
    plot.margin = ggplot2::margin(t = 14, r = 3, b = 6, l = 40)
  )

# Shared legend in a single row
p_legend_source <- p_viterbi_state_prop_by_success +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      nrow = 1,
      byrow = TRUE,
      keywidth = grid::unit(0.6, "cm"),
      keyheight = grid::unit(0.6, "cm")
    )
  ) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.justification = "left",
    legend.text = ggplot2::element_text(size = 15),
    legend.key.size = grid::unit(0.6, "cm"),
    legend.spacing.x = grid::unit(0.15, "cm"),
    legend.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0)
  )

common_legend <- cowplot::get_legend(p_legend_source)

p_viterbi_panels <- cowplot::plot_grid(
  NULL,
  p_cov_panel,
  NULL,
  p_ind_panel,
  NULL,
  ncol = 5,
  align = "h",
  axis = "tb",
  rel_widths = c(0.01, 1.12, 0.06, 1.12, 0.01)
)

# Place the legend at the lower left
legend_left <- cowplot::plot_grid(
  common_legend,
  NULL,
  ncol = 2,
  rel_widths = c(0.72, 1)
)

p_viterbi_state_prop_combined <- cowplot::plot_grid(
  p_viterbi_panels,
  legend_left,
  ncol = 1,
  rel_heights = c(1, 0.10)
)

p_viterbi_state_prop_combined

save_pdf(
  filename = here::here(
    "figure",
    "Viterbi_state_proportion_combined.pdf"
  ),
  plot = p_viterbi_state_prop_combined,
  size = FIGURE_SIZES$viterbi_combined
)
```
