---
title: "Trip-level movement analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(lubridate)
library(here)
library(geosphere)
library(sf)
library(scales)
library(glmmTMB)
library(broom.mixed)
library(patchwork)
library(colorspace)

source(here::here("code", "_shared.R"))
```

```{r directories, include=FALSE}
ensure_project_dirs()
```

```{r settings, include=FALSE}
TRIP_PATHS <- list(
  data = here::here("outputs", "trip_segmentation.csv"),
  breakwater = here::here("rawdata", "Kab_4.gpkg"),
  coastline = here::here("rawdata", "gadm41_JPN_shp", "gadm41_JPN_0.shp"),
  figure = here::here("figure", "S4_trip_parameters_p.pdf")
)

SEA_TRIP_THRESHOLD <- 0.70
SEA_TRIP_THRESHOLDS <- c(0.70, 0.75, 0.80, 0.85, 0.90, 0.95)

assert_input_files(unlist(TRIP_PATHS[c("data", "breakwater", "coastline")]))

trip_figure_theme <- function() {
  list(
    ggplot2::theme_classic(),
    ggplot2::theme(
      axis.title.x = ggplot2::element_text(size = 19, margin = ggplot2::margin(t = 12)),
      axis.title.y = ggplot2::element_text(size = 19, margin = ggplot2::margin(r = 12)),
      axis.text = ggplot2::element_text(size = 16),
      legend.position = "none",
      axis.line = ggplot2::element_blank(),
      panel.border = ggplot2::element_rect(
        colour = "black",
        fill = NA,
        linewidth = 0.8
      )
    )
  )
}
```

# Data input

```{r}
trip_data <- read.csv(TRIP_PATHS$data)
```

Bird-level information

```{r}
bird_info <- tibble::tribble(
  ~id, ~success_rate, ~sex, ~hatching_day,
  "62206", 0.40, "M", "5/23",
  "62207", 0.00, "F", "5/23",
  "671", 0.60, "F", "5/20",
  "675", 0.50, "F", "5/23",
  "681", 0.40, "M", "5/19",
  "683", 0.00, "M", "5/19",
  "685", 0.38, "M", "5/19",
  "ATK_29B_S1", 0.00, "M", "5/18",
  "ATK_2B_S1", 0.40, "M", "5/18",
  "ATK_6E_S1", 0.60, "F", "5/18",
  "BTG_KAB_2025_9B01831_S1", 0.11, "F", "5/23",
  "BTG_KAB_2025_9B03427_S1", 0.67, "F", "5/19",
  "BTG_KAB_2025_9B40989_S2", 0.25, "F", "5/19"
)
```

# Data processing

Classify locations

```{r}
# Breakwater polygons
kab <- sf::st_read(
  TRIP_PATHS$breakwater,
  layer = "kab4__kab",
  quiet = TRUE
) %>%
  dplyr::mutate(
    category = dplyr::if_else(!is.na(breakwater), "breakwater", NA_character_)
  ) %>%
  dplyr::filter(category == "breakwater") %>%
  sf::st_make_valid()

# Coastline polygon
coast <- sf::st_read(
  TRIP_PATHS$coastline,
  quiet = TRUE
) %>%
  sf::st_make_valid()

# GPS points
trip_sf <- sf::st_as_sf(
  trip_data,
  coords = c("lon", "lat"),
  crs = 4326
)

# Land/sea classification
sf::sf_use_s2(TRUE)

trip_located <- sf::st_join(
  trip_sf,
  coast,
  left = TRUE
) %>%
  dplyr::mutate(
    coast_location = dplyr::if_else(is.na(COUNTRY), "sea", "land")
  )

sf::sf_use_s2(FALSE)

trip_located <- trip_located %>%
  dplyr::mutate(
    on_breakwater = lengths(sf::st_intersects(trip_located, kab)) > 0,
    location = dplyr::if_else(on_breakwater, "land", coast_location)
  ) %>%
  dplyr::select(
    -dplyr::any_of(c("GID_0", "COUNTRY", "coast_location"))
  ) %>%
  sf::st_drop_geometry()
```

Select trips with >= 70% sea locations

```{r}
sea_threshold <- SEA_TRIP_THRESHOLD

trip_sea_summary <- trip_located %>%
  dplyr::filter(
    !is.na(Trip_ID),
    !on_breakwater
  ) %>%
  dplyr::group_by(id, Trip_ID) %>%
  dplyr::summarise(
    n_points = dplyr::n(),
    n_sea = sum(location == "sea", na.rm = TRUE),
    prop_sea = n_sea / n_points,
    .groups = "drop"
  )

sea_trip_keys <- trip_sea_summary %>%
  dplyr::filter(prop_sea >= sea_threshold) %>%
  dplyr::select(id, Trip_ID, prop_sea)
```

# GLMMs

Calculate trip-level parameters

```{r}
build_trip_parameters <- function(trip_keys) {
  trip_data %>%
    dplyr::inner_join(
      trip_keys,
      by = c("id", "Trip_ID")
    ) %>%
    dplyr::arrange(id, Trip_ID, time) %>%
    dplyr::group_by(id, Trip_ID) %>%
    dplyr::mutate(
      step_length_m = geosphere::distHaversine(
        cbind(lon, lat),
        cbind(dplyr::lag(lon), dplyr::lag(lat))
      )
    ) %>%
    dplyr::summarise(
      total_distance = sum(step_length_m, na.rm = TRUE) / 1000,
      max_distance = max(Nest_Distance, na.rm = TRUE) / 1000,
      trip_duration = sum(time_diff_sec, na.rm = TRUE) / 60,
      Date = as.Date(min(time)),
      year = dplyr::first(year),
      logger = dplyr::first(logger),
      prop_sea = dplyr::first(prop_sea),
      .groups = "drop"
    ) %>%
    dplyr::left_join(bird_info, by = "id") %>%
    dplyr::mutate(
      hatching_day = paste0(
        stringr::str_pad(
          stringr::str_extract(hatching_day, "^\\d+"),
          width = 2,
          pad = "0"
        ),
        "-",
        stringr::str_pad(
          stringr::str_extract(hatching_day, "\\d+$"),
          width = 2,
          pad = "0"
        )
      ),
      hatch_date = as.Date(paste0(year, "-", hatching_day)),
      days_from_hatching = as.numeric(Date - hatch_date),
      attachment_method = dplyr::case_when(
        logger %in% c("Gipsy", "VHF") ~ "harness",
        logger == "Axy" ~ "tesa",
        TRUE ~ NA_character_
      ),
      id = factor(id),
      sex = factor(sex),
      logger = factor(logger),
      attachment_method = factor(attachment_method),
      Date = factor(Date)
    ) %>%
    dplyr::select(
      id,
      Trip_ID,
      prop_sea,
      Date,
      year,
      logger,
      attachment_method,
      sex,
      success_rate,
      days_from_hatching,
      total_distance,
      max_distance,
      trip_duration
    )
}

data_para <- build_trip_parameters(sea_trip_keys)

data_para
```

Fit trip-level GLMMs

```{r}
fit_trip_model <- function(df, response_var) {
  
  df_model <- df %>%
    dplyr::mutate(
      response_value = .data[[response_var]]
    ) %>%
    dplyr::filter(
      !is.na(response_value),
      response_value > 0,
      !is.na(success_rate),
      !is.na(days_from_hatching),
      !is.na(sex),
      !is.na(id),
      !is.na(Date)
    )
  
  formula_model <- stats::as.formula(
    paste0(
      "response_value ~ success_rate + days_from_hatching + sex + ",
      "(1 | id) + (1 | Date)"
    )
  )
  
  glmmTMB::glmmTMB(
    formula_model,
    data = df_model,
    family = glmmTMB::lognormal(link = "log")
  )
}

response_vars <- c(
  "total_distance",
  "max_distance",
  "trip_duration"
)

count_model_data <- function(df, response_var) {
  
  df %>%
    dplyr::mutate(
      response_value = .data[[response_var]]
    ) %>%
    dplyr::filter(
      !is.na(response_value),
      response_value > 0,
      !is.na(success_rate),
      !is.na(days_from_hatching),
      !is.na(sex),
      !is.na(id),
      !is.na(Date)
    )
}

model_results <- tibble::tibble(
  response_var = response_vars
) %>%
  dplyr::mutate(
    model_data = purrr::map(
      response_var,
      \(rv) count_model_data(data_para, rv)
    ),
    model = purrr::map(
      response_var,
      \(rv) fit_trip_model(data_para, rv)
    ),
    n_trips = purrr::map_int(
      model_data,
      nrow
    ),
    n_ids = purrr::map_int(
      model_data,
      \(df) dplyr::n_distinct(df$id)
    ),
    converged = purrr::map_lgl(
      model,
      \(m) isTRUE(m$sdr$pdHess)
    ),
    AIC = purrr::map_dbl(
      model,
      AIC
    )
  )

model_results
```

Model Summaries

```{r}
model_results %>%
  dplyr::select(response_var, n_trips, n_ids, AIC, converged, model) %>%
  purrr::pwalk(function(response_var, n_trips, n_ids, AIC, converged, model) {
    
    cat("\n\n")
    cat("============================================================\n")
    cat("Response variable:", response_var, "\n")
    cat("n trips          :", n_trips, "\n")
    cat("============================================================\n\n")
    
    print(summary(model))
  })
```

### Figure S4: Trip-level parameters by success rate

```{r}
y_vars <- c(
  "total_distance",
  "max_distance",
  "trip_duration"
)

y_labels <- c(
  total_distance = "Total distance (km)",
  max_distance = "Maximum distance (km)",
  trip_duration = "Trip duration (min)"
)

data_summary_all <- purrr::map_dfr(y_vars, function(y) {
  
  data_para %>%
    dplyr::group_by(id, success_rate) %>%
    dplyr::summarise(
      mean_y = mean(.data[[y]], na.rm = TRUE),
      sd_y = sd(.data[[y]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      variable = y
    )
})

id_levels <- levels(data_para$id)

id_cols <- setNames(
  colorspace::qualitative_hcl(
    length(id_levels),
    palette = "Dark 3"
  ),
  id_levels
)

plot_list <- purrr::imap(y_vars, function(y, i) {
  
  data_summary <- data_summary_all %>%
    dplyr::filter(variable == y)
  
  x_label <- if (i == length(y_vars)) "Success rate" else NULL
  
  ggplot() +
    geom_jitter(
      data = data_para,
      aes(
        x = success_rate,
        y = .data[[y]],
        colour = id
      ),
      alpha = 0.5,
      size = 1.5,
      width = 0.02,
      height = 0
    ) +
    geom_errorbar(
      data = data_summary,
      aes(
        x = success_rate,
        ymin = mean_y - sd_y,
        ymax = mean_y + sd_y,
        colour = id
      ),
      width = 0.05,
      linewidth = 0.8
    ) +
    geom_point(
      data = data_summary,
      aes(
        x = success_rate,
        y = mean_y,
        colour = id
      ),
      size = 4
    ) +
    scale_colour_manual(values = id_cols) +
    trip_figure_theme() +
    labs(
      x = x_label,
      y = y_labels[y]
    )
})

fig_s3 <- patchwork::wrap_plots(
  plot_list,
  ncol = 1
)

fig_s3

save_pdf(
  filename = TRIP_PATHS$figure,
  plot = fig_s3,
  size = FIGURE_SIZES$trip_parameters
)
```





# Data check

GPS data collection days per individual

```{r}
# ------------------------------------------------------------
# GPS tracking duration per individual
# ------------------------------------------------------------

gps_days_by_id <- trip_data %>%
  dplyr::mutate(
    Date = as.Date(time)
  ) %>%
  dplyr::group_by(id) %>%
  dplyr::summarise(
    first_date = min(Date, na.rm = TRUE),
    last_date = max(Date, na.rm = TRUE),
    gps_days = as.numeric(last_date - first_date) + 1,
    .groups = "drop"
  )

gps_days_by_id


gps_days_summary <- gps_days_by_id %>%
  dplyr::summarise(
    n_individuals = dplyr::n_distinct(id),
    mean_days = mean(gps_days, na.rm = TRUE),
    min_days = min(gps_days, na.rm = TRUE),
    max_days = max(gps_days, na.rm = TRUE),
    sd_days = sd(gps_days, na.rm = TRUE)
  )

gps_days_summary
```
Total number of trips and trips per individual

```{r}
all_trip_keys <- trip_data %>%
  dplyr::filter(!is.na(Trip_ID)) %>%
  dplyr::distinct(id, Trip_ID)

n_all_trips <- nrow(all_trip_keys)

n_all_trips


trips_by_id <- all_trip_keys %>%
  dplyr::count(id, name = "n_trips")

trips_by_id


trips_summary <- trips_by_id %>%
  dplyr::summarise(
    n_individuals = dplyr::n_distinct(id),
    total_trips = sum(n_trips, na.rm = TRUE),
    mean_trips = mean(n_trips, na.rm = TRUE),
    min_trips = min(n_trips, na.rm = TRUE),
    max_trips = max(n_trips, na.rm = TRUE),
    sd_trips = sd(n_trips, na.rm = TRUE)
  )

trips_summary
```

Number and percentage of sea trips

```{r}
sea_trip_keys_80 <- sea_trip_keys %>%
  dplyr::distinct(id, Trip_ID)

n_sea_trips <- nrow(sea_trip_keys_80)

percent_sea_trips <- n_sea_trips / n_all_trips * 100

n_sea_trips
percent_sea_trips



sea_trip_summary <- tibble::tibble(
  n_all_trips = n_all_trips,
  n_sea_trips = n_sea_trips,
  percent_sea_trips = percent_sea_trips
)

sea_trip_summary
```
Number of sea trips per individual

```{r}
sea_trips_by_id <- bird_info %>%
  dplyr::select(id) %>%
  dplyr::left_join(
    sea_trip_keys_80 %>%
      dplyr::count(id, name = "n_sea_trips"),
    by = "id"
  ) %>%
  dplyr::mutate(
    n_sea_trips = tidyr::replace_na(n_sea_trips, 0L)
  )

sea_trips_by_id

sea_trips_by_id_summary <- sea_trips_by_id %>%
  dplyr::summarise(
    n_individuals = dplyr::n_distinct(id),
    mean_sea_trips = mean(n_sea_trips, na.rm = TRUE),
    min_sea_trips = min(n_sea_trips, na.rm = TRUE),
    max_sea_trips = max(n_sea_trips, na.rm = TRUE),
    sd_sea_trips = sd(n_sea_trips, na.rm = TRUE)
  )

sea_trips_by_id_summary
```

### Sensitivity analysis for sea-trip threshold

```{r}
# ============================================================
# 70, 75, 80, 85, 90, and 95%
# Extract success-rate effects from the same GLMM structure
# ============================================================

sea_thresholds <- SEA_TRIP_THRESHOLDS

response_vars <- c(
  "total_distance",
  "max_distance",
  "trip_duration"
)

response_labels <- c(
  total_distance = "Total distance",
  max_distance   = "Maximum distance",
  trip_duration  = "Trip duration"
)

# ------------------------------------------------------------
# Function to create trip-level data for one sea-fix threshold
# ------------------------------------------------------------

make_trip_data_threshold <- function(threshold) {
  sea_trip_keys_threshold <- trip_sea_summary %>%
    dplyr::filter(prop_sea >= threshold) %>%
    dplyr::select(id, Trip_ID, prop_sea)

  build_trip_parameters(sea_trip_keys_threshold)
}

# ------------------------------------------------------------
# Function to fit one GLMM and extract success-rate statistics
# ------------------------------------------------------------

fit_extract_success <- function(df, response_var) {

  df_model <- df %>%
    dplyr::mutate(
      response_value = .data[[response_var]]
    ) %>%
    dplyr::filter(
      !is.na(response_value),
      response_value > 0,
      !is.na(success_rate),
      !is.na(days_from_hatching),
      !is.na(sex),
      !is.na(id),
      !is.na(Date)
    )

  formula_model <-
    response_value ~
      success_rate +
      days_from_hatching +
      sex +
      (1 | id) +
      (1 | Date)

  # First fit the model with the default nlminb optimizer
  model_nlminb <- glmmTMB::glmmTMB(
    formula_model,
    data = df_model,
    family = glmmTMB::lognormal(link = "log"),
    control = glmmTMB::glmmTMBControl(
      optimizer = stats::nlminb
    )
  )

  coef_nlminb <- summary(model_nlminb)$coefficients$cond
  se_nlminb <- unname(coef_nlminb["success_rate", "Std. Error"])

  # Keep the nlminb fit when standard errors are available
  if (is.finite(se_nlminb)) {

    model <- model_nlminb
    optimizer_used <- "nlminb"

  } else {

    # Refit with BFGS when standard errors are unavailable
    model_bfgs <- glmmTMB::glmmTMB(
      formula_model,
      data = df_model,
      family = glmmTMB::lognormal(link = "log"),
      control = glmmTMB::glmmTMBControl(
        optimizer = stats::optim,
        optArgs = list(method = "BFGS")
      )
    )

    coef_bfgs <- summary(model_bfgs)$coefficients$cond
    se_bfgs <- unname(coef_bfgs["success_rate", "Std. Error"])

    model <- model_bfgs
    optimizer_used <- "BFGS"
  }

  coef_table <- summary(model)$coefficients$cond
  success_row <- coef_table["success_rate", ]

  tibble::tibble(
    n_trips = nrow(df_model),
    n_ids = dplyr::n_distinct(df_model$id),
    n_dates = dplyr::n_distinct(df_model$Date),
    beta = unname(success_row["Estimate"]),
    SE = unname(success_row["Std. Error"]),
    z = unname(success_row["z value"]),
    p = unname(success_row["Pr(>|z|)"]),
    pdHess = isTRUE(model$sdr$pdHess),
    convergence_code = model$fit$convergence,
    convergence_message = model$fit$message,
    optimizer = optimizer_used,
    model = list(model)
  )
}

# ------------------------------------------------------------
# Run all thresholds and response variables
# ------------------------------------------------------------

sensitivity_results <- tidyr::crossing(
  threshold = sea_thresholds,
  response_var = response_vars
) %>%
  dplyr::mutate(
    threshold_data = purrr::map(
      threshold,
      make_trip_data_threshold
    ),
    result = purrr::map2(
      threshold_data,
      response_var,
      fit_extract_success
    )
  ) %>%
  dplyr::select(
    threshold,
    response_var,
    result
  ) %>%
  tidyr::unnest(result) %>%
  dplyr::mutate(
    response_variable = unname(response_labels[response_var]),
    threshold_percent = threshold * 100,
    response_variable = factor(
      response_variable,
      levels = c(
        "Total distance",
        "Maximum distance",
        "Trip duration"
      )
    )
  ) %>%
  dplyr::arrange(
    response_variable,
    threshold_percent
  )

# Inspect models with convergence problems
problem_models <- sensitivity_results %>%
  dplyr::filter(
    !is.finite(SE) |
      !pdHess |
      convergence_code != 0
  ) %>%
  dplyr::select(
    response_variable,
    threshold_percent,
    beta,
    SE,
    p,
    optimizer,
    pdHess,
    convergence_code,
    convergence_message
  )

problem_models

# ------------------------------------------------------------
# Format table for the Supplementary Materials
# ------------------------------------------------------------

sensitivity_table <- sensitivity_results %>%
  dplyr::mutate(
    beta = sprintf("%.2f", beta),
    SE = sprintf("%.2f", SE),
    z = sprintf("%.2f", z),
    p = dplyr::case_when(
      p < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.2f", p)
    )
  ) %>%
  dplyr::select(
  `Response variable` = response_variable,
  `Threshold (%)` = threshold_percent,
  `n trips` = n_trips,
  `β` = beta,
  `SE` = SE,
  `z` = z,
  `p` = p
) %>%
  dplyr::group_by(`Response variable`) %>%
  dplyr::mutate(
    `Response variable` = as.character(`Response variable`),
    `Response variable` = dplyr::if_else(
      dplyr::row_number() == 1,
      `Response variable`,
      ""
    )
  ) %>%
  dplyr::ungroup()

sensitivity_table
```
